\section{Big data analytics}

\subsection{Introduction}

\subsubsection{What is Big data}

Big data is a buzzword used to characterize certain kind of data. Those data were originally characterized by their volume, velocity and variety, but now we have more dimensions. Big data processing life cycle follows this path

\begin{enumerate}
    \item Acquisition
    \item Preprocessing
    \item Storage and management
    \item Privacy and security
    \item Analyzing
    \item Visualization
\end{enumerate}

Usually these datasets are very huge and traditional tools can't process them efficiently. 

\vspace{10pt}

Traditional data 

\begin{itemize}
    \item Relational
    \item Structured
    \item Manageable in SQL systems
\end{itemize}

Big data
\begin{itemize}
    \item Distributed across systems
    \item Structured, semi-structured or unstructured
    \item Requires new technology
\end{itemize}

With traditional data the best way to manage them is by vertically scaling the requirements, with big data we have to grow horizontally. 

\subsubsection{Data sources}

The majority of big data come from the web revolution with drivers that are everyday more common

\begin{itemize}
    \item Internet
    \item Smartphones
    \item Social media
    \item Sensors
\end{itemize}

The data can be considered by their origins

\begin{itemize}
    \item Human generated
    \item Machine generated
    \item Business and institutional
    \item Open/public
\end{itemize}

\subsubsection{The Vs of Big data}

The 3 Vs

\begin{enumerate}
    \item Volume
    \item Velocity 
    \item Variety 
\end{enumerate}

5 additional Vs
\begin{enumerate}
    \item Veracity
    \item Value
    \item Variability
    \item Validity
    \item Visualization
\end{enumerate}


\subsubsection{The Big data ecosystem}

The place where we process these data, composed of

\begin{itemize}
    \item Storage systems
    \item Process engines
    \item Management and governance
    \item Analytics and visualization tools
    \item People and skills
\end{itemize}

The ecosystem is characterized by 4 layers

\begin{enumerate}
    \item Storage layer
    \item Processing layer
    \item Management layer
    \item Analytics layer
\end{enumerate}

We use many tools to visualise these data and 3 main kinds of people act in the ecosystem

\begin{enumerate}
    \item Data engineers
    \item Data scientists
    \item Business analysts
\end{enumerate}

\subsubsection{Applications of Big data}

Big data analytics as many possible fields of application

\begin{enumerate}
    \item Business and retail
    \item Social media communication
    \item Healthcare and life sciences
    \item Finance and economy
    \item Government and public sector
    \item Smart cities
    \item Science and research
\end{enumerate}


\subsection{Distributed storage, ecosystems and programming paradigms}

\subsubsection{Distributed storage and file systems}

To save our data we could think about saving flat files on local disks, but this comes with problems of redundancy, slow search and retrieval and poor security. Then we can use file systems that define how files are named, stored and retrieved. Those systems use directories and store the data in blocks of given size. 

\vspace{10pt}

As said before we need distribution to save our data in a good way, then we use many servers. To avoid problems given by common node failures we can use replication and sequential writing. The concept of connecting many independent computers to work as one big system is called \textbf{cluster computing}. Each computer runs its own operating system but when combined they form a unified resource. \textbf{Sharding} is another way where we split our big datasets into smaller pieces called shards which are then distributed across multiple machines. 

\vspace{10pt}

Splitting our data let us ensure availability and reliability, usually we have a replication factor of 3. One of the main systems is HDFS (Hadoop Distributed File System). In HDFS we have name nodes, data nodes and replication pipelines among the data nodes. 

\vspace{10pt}

HDFS strengths

\begin{itemize}
    \item Efficient for very large files
    \item Streaming data access
    \item Good on low-cost pieces of hardware
\end{itemize}

HDFS cons
\begin{itemize}
    \item Low latency access
    \item Lots of small files
    \item Multiple writers/edits
\end{itemize}

Another way to store data is using object storage (Cloud). The data are stored as objects in a bucket. Good for elastic scaling but we have higher latency. 

\subsubsection{Distributed computing models}

Divide and conquer approach where we process data where they are stored. We scale horizontally across nodes. 

\subsubsection{The big data ecosystem}

Formed by a web of interconnected tools for storage, processing, analytics and visualization. They will lower costs and time of deployment. 

\vspace{10pt}

Hadoop ecosystem
\begin{itemize}
    \item HDFS for storage
    \item YARN to manage resources
    \item MapReduce for batch processing
    \item Hive, Pig, Sqoop, Flume for analytics
\end{itemize}

Apache Spark

\begin{itemize}
    \item Unifier engine for multiple tools
    \item Faster than MapReduce
    \item Multiple programming languages
\end{itemize}

The databases are usually NoSQL and can be
\begin{itemize}
    \item Key-Value
    \item Document
    \item Columnar
    \item Graph 
\end{itemize}

For unstructured and structured data we use data lakes, warehouses of lakehoues

\subsubsection{Programming paradigms}

As said we need to parallelize and resist faulting when working with big data. We can have imperative (step by step), functional (we define our own functions) and declarative (what to compute, not how) programming. 

\vspace{10pt}

The most common approach is functional programming where computation is considered as evaluation of functions. We use things like map, reduce, filter and fold.

\vspace{10pt}

MapReduce paradigm
\begin{itemize}
    \item Apply function to data blocks (Map step)
    \item Group by Key (Shuffle step)
    \item Aggregate results (Reduce step)
\end{itemize}


\subsection{Distributed parallel processing and the map reduce paradigm}

\subsubsection{Distributed parallel processing}

Using big data we deal with terabytes of data, so we can't put them on a single server. The execution will be too slow, then we will split the workload into smaller tasks that will run simultaneously and possibly in parallel. 

\begin{itemize}
    \item \textbf{Parallel} \ra multiple tasks executed at the same time
    \item \textbf{Concurrent} \ra multiple tasks that progress together but not necessarily at the same time
\end{itemize}

The traditional way of analyzing data is to move all the data to one machine and process them. This is inefficient, to save network bandwidth we can do a distributed approach where we send the computations where the data already is. The architecture is cluster like and made by nodes organized in racks. 

\subsubsection{The MapReduce paradigm}

Paradigm where we try to reduce complexity. We can use Hadoop to resolve logistic problems. 

\textbf{Work flow} 

\begin{enumerate}
    \item \textbf{Map} \ra Process input split and emit key-value pairs
    \item \textbf{Shuffle/Sort} \ra group intermediate pairs by key
    \item \textbf{Reduce } \ra Aggregate values for each key to get a final output
\end{enumerate}

\begin{verbatim}
    def map(key, value):
        for word in value.split():
        emit(word, 1)

    def reduce(key, values):
        emit(key, sum(values))
\end{verbatim}

\subsubsection{Case study}

\begin{itemize}
    \item Temperature records
    \item Retail analytics
\end{itemize}

Strengths of MapReduce
\begin{itemize}
    \item Scales to thousands of nodes
    \item Fault-tolerant
    \item Simplifies distributed programming
\end{itemize}

Limitations
\begin{itemize}
    \item Batch only, causes latency
    \item Heavy disk
    \item Iterative and ML inefficient
\end{itemize}

\subsection{The Hadoop ecosystem}

\subsubsection{Hadoop}

Hadoop is an open source framework for distributed storage and processing of large datasets. It's designed to scale horizontally across clusters.

Key features are fault tolerance (data replication), high throughput (parallel processing) and scalability (from few nodes to thousands). 
Hadoop has two main pillars

\begin{enumerate}
    \item HDFS \ra Hadoop distributed file system (stores data across multiple machines and replicates blocks for reliability)
    \item MapReduce \ra processing engine (splits computation between map and reduce, extended with yarn to manage cluster resources)
\end{enumerate}

Before Hadoop, we had to fit our data into one machine (limited scalability). Hadoop allowed big data storage and parallel computation.

\subsubsection{Filesystems and HDFS basics}

Since large datasets can't fit on one machine we have to distribute it across nodes and to make computation where data resides. 

Terms of Filesystems

\begin{itemize}
    \item \textbf{Block} \ra fixed-sixe unit of storage
    \item \textbf{Split} \ra logical division of input for MapReduce
    \item \textbf{Shard/Partition} terms for splitting datasets
\end{itemize}

Filesystems can be local (one disk, not fault-tolerant) or distributed (replicates blocks across many nodes). HDFS stores large files and is optimized for throughput, write once and read many times.

\vspace{10pt}

HDFS components
\begin{itemize}
    \item NameNode \ra Manages metadata and block locations
    \item Secondary NameNode \ra assist with checkpoints
    \item DataNodes \ra store actual blocks
\end{itemize}

To ensure fault tolerance we have block replication, one block is stored in multiple nodes or racks. Large block size reduces metadata overhead, and we have data locality for the computation. 

\subsubsection{Hadoop's execution model}

Hadoop originally used MapReduce v1 both as computation engine and resource manager, but there were some problems like single job tracker, limited scalability and rigidity. YARN was introduced to fix this in Hadoop 2.0.

\textbf{V1}

\begin{itemize}
    \item We have a JobTracker that assigns and manages tasks.
    \item The TaskTracker runs the map and reduce tasks and report the status to the JobTracker. 
    \item If JobTracker is overloaded it will break.
    \item As the clusters grow we have a bottleneck that will easily break the system.
\end{itemize}

\textbf{YARN}

\begin{itemize}
    \item Separates resource management from job execution.
    \item The key roles are ResourceManager, NodeManager and ApplicationMaster
    \item It supports multiple processing frameworks
    \item Allows for better scalability, reliability and flexibility
\end{itemize}

\textbf{YARN Architecture Components}\begin{itemize}\item \textbf{ResourceManager (RM):} The global authority of the cluster. It handles resource scheduling (via the \textit{Scheduler}) and job submissions (via the \textit{ApplicationManager}). It tracks availability of $\text{CPU}$, memory, and $\text{I/O}$.\item \textbf{NodeManager (NM):} Runs on each worker node. It manages local resources, launches tasks inside \textbf{Containers}, and reports health and usage back to the RM.\item \textbf{ApplicationMaster (AM):} One instance per job. It negotiates resources with the RM and manages the job lifecycle (requesting containers, handling failures) without micromanaging every task.\end{itemize}\textbf{ Containers and Locality}\begin{itemize}\item \textbf{Containers:} A logical bundle of resources ($\text{CPU} + \text{Memory} + \text{I/O}$). Every task runs in its own isolated container to prevent conflicts.\item \textbf{Locality Awareness:} To minimize network costs, YARN prefers \textbf{Node-local} execution (data on the same node), then \textbf{Rack-local}, and finally \textbf{Off-switch} as a last resort.\end{itemize}\textbf{ Scheduling and Benefits}\begin{itemize}\item \textbf{Schedulers:} Supports \textit{FIFO} (simple order), \textit{Capacity} (guaranteed queues), and \textit{Fair} (equal sharing/preemption).\item \textbf{Advantages:} Enables multitenancy, scales to tens of thousands of nodes, and supports multiple engines like Spark, Tez, and Flink beyond just MapReduce.\end{itemize}

\subsubsection{Optimization and refinements}

\textbf{From Splits to Reducers}
\begin{itemize}
    \item \textbf{Input Splits \& Blocks:} Large files are broken into HDFS blocks (default 128 MB). Each split is assigned to a map task; parallelism depends on the number of blocks.
    \item \textbf{Shuffle \& Sort:} Mappers produce (key, value) pairs. The system groups all values with the same key across the cluster. This step is \textbf{network-intensive} and often the bottleneck of MapReduce.
    \item \textbf{Reducers:} Each reducer receives one or more key groups, applies the function (sum, max, etc.), and writes the final output back to HDFS.
\end{itemize}



\textbf{Optimizations: Combiners \& Partitioners}
\begin{itemize}
    \item \textbf{Combiners:} Optional "mini-reducers" on mapper output. They perform local aggregation before the shuffle phase to reduce network traffic (e.g., local word count sums).
    \item \textbf{Partitioners:} Decide which reducer processes a specific key (default: $hash(key) \pmod{numReducers}$). Used for load balancing or specific workflows.
\end{itemize}



\textbf{Handling Stragglers \& Joins}
\begin{itemize}
    \item \textbf{Speculative Execution:} To handle slow nodes ("stragglers"), Hadoop launches duplicate copies of slow tasks on other nodes. The first copy to finish is kept, reducing job completion time.
    \item \textbf{Reduce-side Joins:} Used for joining datasets on a key (e.g., $R(A,B) \bowtie S(B,C)$). Mappers emit (B, row) from both datasets, and the reducer performs the join. Flexible but expensive due to high shuffle overhead.
\end{itemize}

\textbf{Hadoop Strengths \& Limitations}
\begin{itemize}
    \item \textbf{Strengths:} Handles petabytes of data, fault-tolerant (automatic replication/retries), and supported by a large ecosystem (Hive, HBase, Sqoop).
    \item \textbf{Limitations:} \textbf{Batch-oriented} (not suitable for real-time), high latency due to heavy disk I/O, and inefficient for iterative jobs (ML, graph algorithms).
\end{itemize}
\subsubsection{YARN and scheduling}

\textbf{Resource Management in YARN}\begin{itemize}\item \textbf{Resource Types:} YARN manages four key resources across the cluster: $\text{CPU}$ (cores), \textbf{Memory} (RAM), \textbf{Disk I/O} (read/write), and \textbf{Network I/O} (bandwidth).\item \textbf{Container Allocation:} Resources are bundled into \textbf{Containers}. Each container represents a specific amount of $\text{CPU}$ and Memory, ensuring jobs get exactly what they requested.\item \textbf{Benefits:} This provides strict \textbf{isolation} between jobs and increases efficiency by sizing containers to actual application needs.\end{itemize}\textbf{ YARN Scheduling Policies}\begin{itemize}\item \textbf{FIFO Scheduler:} Jobs run in the strict order of submission. While simple, it causes head-of-line blocking where small jobs must wait for large ones to finish.\item \textbf{Capacity Scheduler:} The cluster is divided into \textbf{queues} with guaranteed resources (e.g., Department A gets $50\%$ quota). This ensures multi-tenant access, though unused capacity can be "borrowed" by other queues.\item \textbf{Fair Scheduler:} Resources are dynamically shared so that all jobs get a fair share over time. Small interactive jobs can start immediately even if a large job is already running.\end{itemize}\textbf{ Scheduling Scenarios in Practice}\begin{itemize}\item \textbf{Scenario 1 (FIFO):} A massive data-cleaning job starts first; a subsequent small query is blocked until the first job completes, leading to poor responsiveness.\item \textbf{Scenario 2 (Capacity):} A department is guaranteed $50\%$ of the cluster. Even if other departments are busy, the quota ensures they always have access to resources.\item \textbf{Scenario 3 (Fair):} When a small job arrives during a large job's execution, the scheduler reassigns some containers, so both can progress simultaneously.\end{itemize}

\subsubsection{Hive, Pig and Sqoop}


\textbf{Apache Hive: Data Warehouse on Hadoop}
\begin{itemize}
    \item \textbf{Overview:} A data warehouse infrastructure built on top of Hadoop. It provides \textbf{HiveQL} (a SQL-like language), making Hadoop accessible to analysts without Java/MapReduce knowledge.
    \item \textbf{Schema-on-Read:} Unlike traditional databases, Hive loads data into HDFS "as is" and applies the schema only when the data is read. This offers great flexibility for semi-structured data.
    
    
    
    \item \textbf{Architecture:}
    \begin{itemize}
        \item \textbf{Metastore:} Stores table metadata and schemas.
        \item \textbf{Compiler \& Optimizer:} Parses HiveQL and improves the execution plan (e.g., predicate pushdown).
        \item \textbf{Execution Engine:} Converts the plan into MapReduce, Tez, or Spark jobs.
    \end{itemize}
\end{itemize}

\textbf{Hive Optimization \& File Formats}
\begin{itemize}
    \item \textbf{Partitioning:} Splits tables into HDFS directories based on a column (e.g., \textit{year=2024}). It improves performance via \textbf{partition pruning}.
    \item \textbf{Bucketing:} Further divides data based on a hash function to ensure balanced distribution for joins.
    
    
    
    \item \textbf{Storage Formats:}
    \begin{itemize}
        \item \textbf{ORC / Parquet:} Columnar formats, highly compressed, best for analytical queries.
        \item \textbf{Avro:} Row-based format, excellent for schema evolution and streaming.
    \end{itemize}
\end{itemize}

\textbf{Apache Pig: Scripting for ETL}
\begin{itemize}
    \item \textbf{Overview:} A high-level platform for analyzing large data sets using a procedural language called \textbf{Pig Latin}.
    \item \textbf{Use Case:} Designed for complex data pipelines and ETL (Extract, Transform, Load) workflows. It is more "step-by-step" compared to Hive's declarative SQL approach.
    \item \textbf{Example Flow:} \texttt{LOAD} (bring data) $\rightarrow$ \texttt{GROUP} (by key) $\rightarrow$ \texttt{FOREACH} (apply aggregation).
\end{itemize}

\textbf{Apache Sqoop: SQL-to-Hadoop}
\begin{itemize}
    \item \textbf{Overview:} A tool designed for efficiently transferring bulk data between \textbf{Relational Databases} (RDBMS) and the Hadoop ecosystem (HDFS, Hive, or HBase).
    \item \textbf{Functionality:} It uses Map-only jobs to import/export data in parallel via JDBC.
    
    
    
    \item \textbf{Example Command:}
\begin{verbatim}
sqoop import --connect jdbc:mysql://localhost/db --table orders
\end{verbatim}
    \item \textbf{Pros/Cons:} Simple and supports incremental loads, but it is better suited for bulk transfers rather than real-time streaming.
\end{itemize}


\subsubsection{Summary}

\begin{itemize}
    \item \textbf{Execution model} \ra How Hadoop runs MapReduce jobs
    \item \textbf{Refinements} \ra optimizations via combiners, partitioners, speculative executions
    \item \textbf{Ecosystem} \ra Hive, Pig and Sqoop
\end{itemize}

Hadoop it's still foundational but has been partially replaced by Spark for faster analytics












\subsection{Introduction to Spark}

Spark is a unified engine for big data processing. Some advantages of Spark are:
\begin{enumerate}
    \item We can process in parallel large datasets across a cluster
    \item We can perform ad hoc interactive queries and visualize datasets. We can also do machine learning models using MLlib
    \item We can implement end-to-end data pipelines from different streams
    \item We can analyze graph data sets and social networks
\end{enumerate}

The success of Spark comes from the fact that is open source, fast and provides parallel and distributed computing. The interface for Python is called PySpark.

Some improvements of Spark with respect to Hadoop
\begin{itemize}
    \item MapReduce on Hadoop is slow
    \item Spark is in-memory so faster and more flexible
    \item Spark is a unified engine for batch, SQL etc.
    \item Spark decouples computation from storage. Spark runs independently of HDFS and keeps data in memory between stages, avoiding repeated disk reads/writes
    \item Spark enables interactive queries and iterative algorithms that are difficult in MapReduce
\end{itemize}

Spark in the big data landscape complements Hadoop, works with different kind of datasets, it has cloud-native integrations, bridges between different data platforms and evolved after Hadoop improving it.

\vspace{10pt}

Spark supports multiple languages like

\begin{itemize}
    \item Scala
    \item Python
    \item Java
    \item R
    \item More via 3rd party APIs
\end{itemize}

All languages run on the same Spark core engine ensuring consistency and scalability. The choice on what to use depends on skills, integration needs and project goals. 

\vspace{10pt}

Spark provides two levels of APIs for working with data

\begin{itemize}
    \item Low-level (RDDs) \ra fine-grained control, defines transformations and actions
    \item High-level (DataFrame and Dataset) \ra structured, optimized queries using Catalyst optimizer and Tungsten engine
\end{itemize}

RDDs are ideal for unstructured or experimental workloads while DF and DS are better for performance and interoperability. Users can mix both depending on their data workflow. 

Some useful Spark applications are Batch ETL, interactive SQL queries, streaming analytics, ML at scale and graph processing.

\vspace{10pt}

The main Spark architecture is made of the followings
\begin{itemize}
    \item Driver program \ra orchestrates the operations (runs SparkContext, defined RDD transformations and actions and collects results)
    \item Cluster manager \ra YARN, Kubernetes and Mesos
    \item Executors \ra run tasks and store data (worker processes on cluster nodes, executes tasks in parallel and cache data in memory for reuse)
\end{itemize}

SparkContext is an old entry point for Spark application, it creates RDDs and manages jobs. It was introduced in early Spark, and it is the core connection to the cluster. This by communicating with the cluster manager to allocate resources. The RDD API depends on the SparkContext for all operations.

\vspace{10pt}

SparkSession is a higher-level entry point that unifies SQL and RDDs APIs.

\vspace{10pt}

In Spark, we have two main operations
\begin{enumerate}
    \item Transformations \ra define a new RDD based on the current one(s). The data inside are never changed, but we want to modify them as needed. These transformations are lazy (they won't be executed until action). Some common transformations are map, filter and flatMap.
    \item Actions \ra return values to the driver or save to storage. Some actions are count, take, collect and SaveAsTextFile
\end{enumerate}

The evaluation is lazy since transformations build a plan while actions trigger the execution.

\vspace{10pt}

\textbf{RDD} \ra Resilient distributed dataset, it's an immutable, distributed collection of objects built via transformations. RDD properties are immutability, lazy evaluation, fault-tolerance and partitions across cluster. We can create RDDs from collections and data in memory, from external storage and from transforming existing RDDs.

\vspace{10pt}

Spark has two main level of execution

\begin{enumerate}
    \item Logical \ra what operations to perform, built when RDD operations are defined, represented as a DAG of transformations
    \item Physical \ra how Spark executes the plan, it splits the DAG into stages and tasks, the DAG scheduler and Task scheduler optimize and distribute the work
\end{enumerate}


\vspace{10pt}

Lazy evaluation follows this procedure

\begin{itemize}
    \item Spark builds DAG of operations
    \item Optimizes before execution
    \item Saves redundant computation
\end{itemize}

Creations and transformations do nothing on their own. They are applied when an action is launched. 

\vspace{10pt}

Transformations on RDD can be by Type
\begin{itemize}
    \item Unary \ra operate on one RDD and obtain one new RDD, usually for element-wise operations like filter, map and flatMap
    \item Binary \ra combine two RDDs, like union, intersection and subtraction
    \item Pair \ra work on (key, value) pairs, like groupbykey, reducebykey and sortbykey
\end{itemize}

We can also define narrow and wide transformations based of the execution implications

\begin{itemize}
    \item Narrow \ra each output partition depends on one input partition
    \item Wide \ra the output depends on multiple input partitions, they trigger shuffles (expensive operations)
\end{itemize}


For ML algorithms some important elements are persist() and cache() because they store RDD in memory. 

\vspace{10pt}

To ensure fault tolerance RDDs follow a lineage graph, so if the partition is lost we can recompute from lineage. It's a built-in recovery mechanism that uses a DAG that records how an RDD was created. We don't store data but we recompute them.

\vspace{10pt}

Different from Spark (in-memory, DAG scheduler and faster) MapReduce is disk-based, simpler and robust.

\vspace{10pt}

The current system supports several cluster managers like Standalone mode, YARN and Kubernetes.

\vspace{10pt}

RDDs are limited by some elements 
\begin{itemize}
    \item Verbose API
    \item No schema
    \item Less optimized than DataFrames
\end{itemize}

\subsection{DataFrames in Spark and Spark-Hadoop relationship}


\subsubsection{Recap}

In the previous section we saw the concept of RDDs that are distributed and fault-tolerant collections. But they are also limited by the fact that they have no schema (so they are inefficient for queries) and that the API is verbose. Our next step will be to review DataFrames since they are structured and optimized (SQL-like).

\subsubsection{DataFrames}

A dataframe is a distributed table-like structure with rows and named columns. Pandas DF and SQL tables inspired them. Each column has a defined type. These DataFrames are built on top of RDDs, but they are faster (in-memory) and easier (table-like).

\vspace{10pt}

With an RDD we tell Spark how to aggregate, while with a dataframe we communicate what we want and Spark will optimize it. 

\vspace{10pt}

On top of that DataFrames have the following benefits

\begin{itemize}
    \item Easy to use
    \item Performance are good
    \item Interoperability
    \item Integration
\end{itemize}

To create a DataFrame we can start from

\begin{itemize}
    \item Structured files
    \item RDDs with schema
    \item External database
\end{itemize}

The schema can be inferred by Spark or defined by the user.

\vspace{10pt}

We recall transformations (return a new dataframe) vs actions (trigger execution). They follow the schema Transformations \ra DAG \ra Actions.

\vspace{10pt}

Some very useful functions are called Window functions \ra They enable analytics like top-N per group.

\vspace{10pt}

The queries are optimized by Catalyst in the following stages

\begin{enumerate}
    \item Analysis
    \item Logical optimization
    \item Physical planning
    \item Code generation
\end{enumerate}

After a query is written Spark builds a logical plan that describes what operations need to be done. We then get the unresolved logical plan that includes references Spark has not yet verified. Catalyst then analyzes the plan using the schema and the metadata creating an analyzed logical plan. This last plan is independent of any physical execution strategy.

\vspace{10pt}

The optimization phase follows these steps

\begin{enumerate}
    \item The Catalyst Optimizer transforms the analyzed logical plan into an optimized logical plan
    \item It applies both rule-based and cost-based optimization
    \item Predicate pushdown (move filters close to the data source)
    \item Projection pruning (remove unnecessary columns)
    \item Constant folding (simplify expressions)
    \item Null elimination and common subexpression removal
    \item The result is an optimized logical plan that preserves semantics but reduces computation
\end{enumerate}

The physical plan defines how Spark will perform the computation
\begin{itemize}
    \item Catalyst's planner takes the optimized logical plan and selects the most efficient execution strategy (SortMergeJoin, BroadcastHashJoin, ShuffleHashJoin, whole-stage code generation, task parallelization and stage pipelining)
    \item We can have multiple candidate physical plans and Spark will choose the best one
    \item Tungsten engine receives the plan, and it will execute it managing the memory 
\end{itemize}

Tungsten executes the physical plan using whole-stage code generation and off-heap memory management

\begin{itemize}
    \item Memory + CPU optimizations
    \item Off-heap memory management
    \item Whole-stage code generation
    \item Vectorized processing
\end{itemize}

Catalyst \ra handles query analysis and optimization ensuring that Spark finds efficient execution strategies automatically

\vspace{10pt}

Tungsten \ra handles execution and performance optimizing CPU and memory usage for physical plan execution.

\vspace{10pt}

These two together form the Spark SQL engine. This engine makes Spark efficient, declarative and adaptable. Spark can be integrated with SQL providing the ability of working with different types of data and formats.

\vspace{10pt}

In this setting the DataFrames are the API and Spark SQL is the interface.

\vspace{10pt}

Performance tips

\begin{itemize}
    \item Use Parquet/ORC instead of CSV
    \item Partition data by common query keys
    \item Cache DataFrames when used
    \item Broadcast joins for small tables
\end{itemize}

From this we can have the takeaway that DataFrames are preferable to RDDs unless RDD operations are needed.

\vspace{10pt}

Use RDDs if
\begin{itemize}
    \item You need low level transformations
    \item You work with unstructured or complex data types
    \item You require fine-grained control over partitions or persistence
\end{itemize}

Use DataFrames when

\begin{itemize}
    \item Data is structured
    \item You want automatic optimization
    \item Concise SQL-style queries are preferable
\end{itemize}

For Python users DataFrames are better. They can resort to RDDs if they need more control.




\subsubsection{Datasets}

Spark has also the option of using datasets, a unified API that combines RDDs and DataFrames. Some benefits are that datasets are type-safe, support both functional and relational operations, available in Scala and Java (not Python) and we can have optimization via Catalyst and Tungsten.

\subsubsection{Spark-Hadoop relationship}


Spark integrates with Hadoop but doesn't depend on them. It can also run standalone or in cloud environments. The read and write come from HDFS, YARN is for resource management, Hive Metastore for table metadata and SQL queries, MapReduce is replaced for in-memory executions.

\vspace{10pt}

The integration is done
\begin{itemize}
    \item On YARN
    \item With Hive
    \item With HDFS
\end{itemize}


\subsection{Spark in Cloud-based computational environments}

The main use of Cloud computing in the context of big data is given by the fact that is elastic and can be used with on-demand options. It's the ideal way to work with distributed data processing. It also removes hardware management overload and some models support pay-as-you-go models. 

\vspace{10pt}

Running Spark in the cloud is also useful because pre-existent clusters can be costly, too rigid, and they require skilled admins. With cloud based nodes on the other hand with more elastic and scalable options. Basically we don't need to manage our clusters manually, we can scale easily, the integration with cloud-native storage is easier, and we can access prebuilt Spark distributions.

\vspace{10pt}

Some benefits of this approach are

\begin{itemize}
    \item The deployment is easy and so the management
    \item We can integrate everything with cloud-native tools
    \item Team collaboration is optimized
\end{itemize}

Some cloud deployment options are

\begin{itemize}
    \item Managed Spark services
    \item Spark on Kubernetes or standalone clusters
    \item Interactive notebooks
\end{itemize}

Spark can be managed in two main ways: Managed and Self-Managed

\vspace{10pt}

Managed Spark

\begin{itemize}
    \item Automatically handles setup, scaling and monitoring
    \item Best for production and teaching
\end{itemize}

\vspace{10pt}

Self-Managed Spark

\begin{itemize}
    \item Full control over configuration and cost
    \item Best for research and experimentation
\end{itemize}

\subsubsection{Databricks}

Databricks is a unified platform for data engineering, analytics and AI. It was created by John Spark (the inventor of Spark). This platform is able to fully manage Spark clusters with a notebook based environment. Databricks is formed by the following components:

\begin{itemize}
    \item Workspace \ra collaborative notebook environment
    \item Clusters \ra auto-scaling Spark clusters
    \item Jobs \ra automated workflows
    \item DBFS \ra Databricks File System (built on cloud storage)
\end{itemize}

Databricks has the following advantages

\begin{itemize}
    \item One-click cluster creation
    \item Integrated visualization and SQL interface
    \item Optimized Spark runtime
    \item Supports all major clouds
\end{itemize}

Databricks follows this workflow
\begin{enumerate}
    \item Upload data to the cloud
    \item Create the Databricks notebook
    \item Initialize Spark session
    \item Read, transform and visualise data
    \item Export the results or train models
\end{enumerate}

Simplifying to \textbf{Raw} \ra \textbf{Clean} \ra \textbf{Analytics}.

\vspace{10pt}

Databricks can be used with its free edition (not for production tho).



\subsubsection{AWS Elastic MapReduce (EMR)}

AWS EMR is an Amazon-managed Hadoop and Spark platform that uses EC2 instances and S3 for storage. These two provide resizable virtual services and object storage services. 

\vspace{10pt}

EMR clusters are made of

\begin{itemize}
    \item Master node \ra job coordination
    \item Core nodes \ra run HDFS and Spark executors
    \item Task nodes \ra compute-only 
\end{itemize}

On EMR Spark environments are pre-configured, and they can be easily integrated with S3. EMRFS allows seamless S3 access.

\vspace{10pt}

The workflow is the following
\begin{enumerate}
    \item Store data in Amazon S3
    \item Launch EMR with Spark application
    \item Process the data and write the results back to S3 or RedShift (data warehouse)
\end{enumerate}

S3 \ra EMR \ra S3 \ra RedShift

\vspace{10pt}

EMR has the following advantages

\begin{itemize}
    \item Flexible cluster customization
    \item Cost-control via auto-termination
    \item Deep integration with AWS analytics tools like Athena and Glue
\end{itemize}


\subsubsection{Google Cloud Dataproc}

Dataproc is a tool that help us manage Hadoop and Spark services on Google Cloud. It can be integrated with services like GCS (data lake), Google BigQuery (Analytics) and Vertex AI (unified ML platforms).

Dataproc is convenient since it help us for fast cluster startup, pay-per-second billing is enabled and some tools like Jupyter, Spark and Hadoop are already installed.

\vspace{10pt}

Dataproc uses compute engine VMs for nodes, the storage is done via GCS (not HDFS) and Job orchestration is done via Cloud Composer (Airflow).



\subsubsection{Azure HDInsight}

HDInsight is a Microsoft's managed big data service. It supports Hadoop, Spark, Hive, Kafka and HBase and has a tight integration with Azure Data Lake. The key features of HDInsight are its good security, the fact that can be easily scaled through Azure and that it can be integrated with Power BI and Synapse.

\vspace{10pt}

A typical use include the following steps

\begin{enumerate}
    \item Data from Azure Data Lake
    \item Spark job on HDInsight
    \item Visualization in Power BI
\end{enumerate}


HDInsight has some pros like a very good Microsoft integration (how is this a pro?) and the cluster management. On the other hand the startup is slow and Databricks is cheaper. 

Some replacements that we can find in Azure are Databricks, Fabric, Azure Synapse Analytics, Azure Data Factory, HDInsight on AKS, Azure Stream Analytics and Azure Data Explorer (Kusto).



\subsubsection{Kubernetes}

Since Spark can run natively on K8s clusters we have some advantages if we want to use Kubernetes. Some of them are that we don't care about which cloud service we use, we have fine-graned container scheduling, and we use easily hybrids of cloud services. But we require some DevOps skills, and we have to manually tuning our resources. 


\subsubsection{Google Colab}


Colab is a free, simple and web-based environment that is very useful for learning and experimenting. For storage, it can be connected to Google Drive. Spark can be installed easily on Colab.

\vspace{10pt}

Some problems with Colab
\begin{itemize}
    \item Not suitable for large datasets
    \item No persistent cluster
    \item Good for demonstrations or teaching only
    \item For production we need dedicated or enterprise servers.
\end{itemize}


\subsubsection{Lightning AI}

Lightning AI is a cloud platform for distributed AI and data workflows. It can be integrated with Spark for large-scale preprocessing. It's focussed both on ML pipelines and not just ETL. In this integration Spark handles data ingestion and transformation while Lighting handles model training and orchestration. It's very suitable for AI labs and teaching advanced pipelines. 

\vspace{10pt}

Load data from S3 or GCS \ra Use Spark to preprocess \ra Lightning AI trains model on cleaned data

\vspace{10pt}

Lightning AI supports GPU acceleration, auto-scaling and monitoring and Jupyter-like notebooks



\subsection{Performance and data engineering in Spark}

\subsubsection{Why Spark jobs are slow}

Different ways of running a query can have different results in terms of execution cost. Useless aggregations can result in worse performance, the use of filters can be advisable. In Spark moving data is more expensive than computing on it and by this Shuffle is the main bottleneck in terms of performance.

\vspace{10pt}

The golden rule in Spark to improve efficiency is to reduce data and movement, this can be achieved by

\begin{itemize}
    \item Filtering early
    \item Select only needed columns
    \item Avoid unnecessary shuffle
\end{itemize}


\subsubsection{How Spark optimizes queries}

Spark does not optimize the queries as written instead it analyzes the query, it rewrites and chooses the most efficient execution. Every Spark job follows the pipeline

\begin{enumerate}
    \item Query
    \item Plan
    \item Optimize
    \item Execute
\end{enumerate}

Usually Spark tries to reduce rows early, reduce columns, minimize shuffle with the goal of reducing data movement. For the Join Spark chooses Broadcast Join or Shuffle Join.



\subsubsection{Data engineering for performance}

Good performance is obtained by following these principles

\begin{enumerate}
    \item Reduce data size
    \item Avoid shuffle
    \item Use efficient formats
    \item Reuse computation
    \item Use correct join strategy
\end{enumerate}

One good format is Parquet, different from CSV because is columnar, compressed and fast. Data can be organized by key and by partition pruning.

Another useful element is caching, we save some results in the memory to reuse them. We should avoid caching if we are in a one-time use setting, data are very large and the memory is limited. 

\vspace{10pt}

Joins are expensive operations, and we can have two types of them

\begin{itemize}
    \item Broadcast \ra Small table sent to all nodes, large table stays distributed
    \item Shuffle \ra Large tables require redistribution, expensive because data move across nodes
\end{itemize}

If we apply repartition we have shuffling, with coalesce we don't.

\vspace{10pt}

Uneven data distribution can also be problematic since if one node is overloaded we can have a bottleneck. To handle this case

\begin{itemize}
    \item Repartition
    \item Filtering early
    \item Key salting
\end{itemize}

The size of the files has an impact too, too many small files means overhead while too large files means poor parallelism.

\vspace{10pt}

The ETL pipeline 

\begin{enumerate}
    \item Raw
    \item Clean
    \item Transform
    \item Store
    \item Query
\end{enumerate}

Real pipeline example

\begin{enumerate}
    \item S3
    \item Spark
    \item Parquet
    \item BI
\end{enumerate}

The processing of the data can be periodic (Batch) or in real-time (streaming).

\vspace{10pt}

Performance checklist

\begin{itemize}
    \item Use Parquet
    \item Filter early
    \item Select needed columns
    \item Avoid shuffle
    \item Use caching wisely
\end{itemize}

Common mistakes

\begin{itemize}
    \item SELECT *
    \item Too many joins
    \item No partitioning
    \item Ignoring skew
\end{itemize}


\subsection{Big data storage technologies and relation with Spark}


\subsubsection{Intro to big data storage}

Given the scale and diversity of big data we can't use traditional relational databases. The problems come from the volume, the velocity and the variety of this kind of data. We need a scalable, distributed and flexible way to store our data. Most modern solutions employ horizontal scalability, fault tolerance, flexible schema and integration with different engines like Spark and Hadoop. 

\vspace{10pt}

Recall that HDFS splits the data in block of 128 MB across nodes, and it's ideal for large sequential reads (batch analysis) but not for small and real-time queries. NoSQL technologies arose to handle these new workloads, we have the main 4 families:

\begin{enumerate}
    \item Key-value Stores
    \item Document Stores
    \item Column-family store
    \item Graph databases
\end{enumerate}

Recall the CAP theorem, BASE and ACID principles from storing and retrieving. 

\vspace{10pt}

Since no single database fits every use case modern architectures combine multiple storage types, this approach is called polyglot persistence. 


\subsubsection{Key-value DB}


This is the simplest form of a NoSQL database. The data are stored as a pair key:value. It's optimized for constant-time lookups, it's schema-free and extremely fast. Some common systems are Redis, Amazon DynamoDB and Riak. They key must be unique, and the values can be of different types. Redis keeps the data in memory, so the operations will be very fast. 

Some operations are 

\begin{itemize}
    \item SET key value \ra store data
    \item GET key \ra retrieve
    \item DEL key \ra delete
    \item INCR key \ra increment counters
    \item EXPIRE key seconds \ra set time-to-live 
\end{itemize}


Strengths
\begin{itemize}
    \item Extremely fast reads/writes
    \item Simple data access model
    \item Scales horizontally
\end{itemize}

Limitations
\begin{itemize}
    \item No complex queries and joins
    \item Limited analytical capability
\end{itemize}

The best use cases are caching, gaming and high-traffic web apps. This approach generalizes the concept of Associative Keys to databases.




\subsubsection{Document DB - MongoDB focus}

This kind of databases store data as JSON-like documents. The schema is flexible, and we can easily query using fields and indexes. Some examples are MongoDB and Couchbase. It's ideal for semi-structured data such as catalogs, users and logs. MongoDB is the most popular document database. It's open-source, widely adopted and stores data and BSON (binary JSON). It balances performance, flexibility and scalability. Used by organizations like eBay, Forbes and CERN. 

\vspace{10pt}

The hierarchy is the following
\begin{enumerate}
    \item Database
    \item Collections
    \item Documents
\end{enumerate}

Even though the schema is not fixed we can still validate it. Every MongoDB document needs to have a '\_id'.

\vspace{10pt}

The usual operations are CRUD (create, read, update and delete)
\begin{enumerate}
    \item Insert
    \item Read
    \item Update
    \item Delete
\end{enumerate}

Using indexes we can improve the query performance. Indexes can be
\begin{itemize}
    \item Single-field
    \item Compound
    \item Text and geo-indexes
\end{itemize}

The aggregation in MongoDB works like a pipeline with some common stages like \$match, \$group, \$sort and \$project.

\vspace{10pt}

Replication in MongoDB ensures data redundancy and fault tolerance by copying data across multiple servers, called replica sets. A replica is made of primary node (handles all write operations) and secondary node (replicate data from the primary asynchronously). If the primary fails, one secondary becomes the new primary (automatic failover) and the clients continue reading from any node. The benefits of this approach are high availability, fault tolerance, data redundancy and read scalability. 

\vspace{10pt}

Sharding is another tool where data are distributed across multiple servers (shards) so that no single node has to handle all the load or store all the data. 
\begin{enumerate}
    \item Data is split into chunks based on a shard key
    \item Each chunk is stored on a different shard
    \item A config server keeps track of where each chunk resides
    \item A mongo's router direct client queries to the appropriate shard
\end{enumerate}

The benefits are horizontal scalability, balance of read and write load across nodes and geographical distribution is supported.

\vspace{10pt}

In production, we usually combine replication ad sharding to make global scalability possible. Each shard is a replica set, this provides both scalability and fault tolerance. 

\vspace{10pt}

MongoDB has some strengths (flexible schema, JSON integration, great for rapid prototyping and rich aggregation features) and limitations (joins are limited, ACID are weaker, and we require index management)

\vspace{10pt}

The main ecosystems related to MongoDB are 
\begin{itemize}
    \item MongoDB Atlas
    \item Compass
    \item BI Connector
    \item Spark Connector
\end{itemize}


\subsubsection{Wide column DB}

This typology of database store data in rows and dynamic columns grouped into column families. It's inspired by Google's Bigtable. Some examples are Apache Cassandra and Apache HBase. The first one doesn't rely on Hadoop while the second one does. Wide columns DB are best for high-write, time-series and IoT data. 

The table structure is like relational databases but with more flexibility and scalability. Each row will have a different set of columns, and columns are grouped into column families rather than fixed tables (like a nested key-value store). Data is stored by column rather than by row. This structure is optimized for aggregations on specific columns. The usual Cassandra operations are Create, Insert, Query and Delete. 

\vspace{10pt}

As always we have clusters (here peer-to-peer with no master node), data is replicated across nodes for fault tolerance, and we have tunable consistency (one, quorum and all). We can see Wide columns DBs as a collection of spreadsheets where each row have its own set of columns depending on what data it has. 

\vspace{10pt}

Some design guidelines are 
\begin{itemize}
    \item Denormalize instead of join
    \item Use both column names and column values to store data
    \item Model an entity with a single row
\end{itemize}


\subsubsection{Graph DB}

Graph databases store data as nodes, edges and properties in order to traverse relationships easily. Some examples are Neo4j, JanusGraph and TigerGraph. With this structure queries can be fast and intuitive. Some easy common use cases are social networks, fraud detection, knowledge graphs and recommendation systems. 


\subsubsection{Other DB types} 

We also have additional kinds of databases like
\begin{itemize}
    \item Search databases \ra Optimized for text search and analytics. Some examples are Elasticsearch and Apache Solr. Inverted indexes are used for fast keywords and full-text queries. These are great for log analytics, monitoring and search engines.
    \item Time-series databases \ra Specialized for sequential data. Some examples are InfluxDB, TimescaleDB and Prometheus. This structure is optimized for regular insertions and aggregations over time windows
    \item Vector databases \ra New generation databases for AI and similarity search. Some examples are Pinecone, Weaviate, Milvus and FAISS. The database stores embeddings of our numerical vectors and can be used for semantic search, recommendation and image/text similarity.
\end{itemize}















