\section{Computational intelligence}

\subsection{Introduction}

Definitions to start the course 
\begin{itemize}
    \item \textbf{Problem} \ra Task to be solved automatically (find the maximum number in a list)
    \item \textbf{Algorithm} \ra Process to solve a problem, set of precise and not ambiguous actions that all together will allow us to reach the solution
    \item \textbf{Program} \ra Translation of an algorithm into a programming language
\end{itemize}

Motivation

\begin{itemize}
    \item In classical computational methods we start from a problem, and we have a human being that imagines an algorithm (problem-solving, it's creative and informal). Then software engineers will put the ideas into a computer in order to get a program (implementation, it's automatic and not creative). 
    \item We need computational intelligence because traditional methods in many cases fail, usually on the first step. We can have many problems where imaging solutions or algorithms it's either impossible or very hard for human being to imagine a solution. This is the case of \textbf{complex problems}
\end{itemize}


Some examples of complex problems
\begin{itemize}
    \item Categorization of clients
    \item Face recognition
    \item Find a new drug to cure a disease
    \item Driving a robot in a 3-D space 
\end{itemize}

Our idea is to give computers or informatic systems the ability to learn autonomously how to solve a problem. We define \textbf{learning} as improving something by means of experience. Those computers can also understand that the generated solution in not good enough, and so they should understand a better way to find the optimal solution. 

\vspace{10pt}

One of the most famous ways to improve those computers are bio-inspired algorithms. Like the \textbf{neural networks} (took inspiration from the brain), \textbf{evolutionary computation} (theory of evolution from Darwin, considering evolution as a sort of learning) where we have genetic algorithms, differential evolution or genetic programming, \textbf{swarm intelligence based algorithms} (system of animals are better at solve problems with respect to singular entities) like particle swarm optimization, \textbf{fuzzy systems} and \textbf{local search} like hill climbing simulated annealing.  

\vspace{10pt}

One category of complex problems are the so-called optimization problems. Even though they are complex they are easy to comprehend and to analyze.

Informally solving an optimization problem means to find the best solution(s) in a typically huge set of possible alternative solutions. We are able to find if something is a solution and to evaluate the quality of the solution with respect to the others. More formally an optimization problem is a pair of objects S and f where
\begin{itemize}
    \item S \ra set of all existing solutions (search space)
    \item f \ra function that evaluates the solutions (fitness or cost or quality function)
\end{itemize}

An optimization problem can be:

\[
\textbf{Maximization:} \quad \text{find } x^\ast \in S \text{ such that } 
f(x^\ast) \geq f(y) \quad \forall y \in S.
\]

\[
\textbf{Minimization:} \quad \text{find } x^\ast \in S \text{ such that } 
f(x^\ast) \leq f(y) \quad \forall y \in S.
\]

To solve this kind of problems we can use an optimization algorithm, a machine that is able to output multiple solutions (iterative machine). 

\vspace{10pt}

\textbf{No free lunch theorem} \ra the average performance of all optimization algorithms calculated on all existing optimization problems is the same. We care about the probability of finding the best solution and the time to reach it. We can't have one algorithm for every problem then we need to find a solution for every problem.


\subsection{Optimization algorithms}

\textbf{Optimization problem} \ra a pair (S, f) where

\begin{itemize}
    \item S is the set of all solutions (search space)
    \item f is the function (fitness function) where f : S \ra \begin{math}
        \mathbb{R} 
    \end{math}
\end{itemize}

An optimization problem can be:

\[
\textbf{Maximization:} \quad \text{find } x^\ast \in S \text{ such that } 
f(x^\ast) \geq f(y) \quad \forall y \in S.
\]

\[
\textbf{Minimization:} \quad \text{find } x^\ast \in S \text{ such that } 
f(x^\ast) \leq f(y) \quad \forall y \in S.
\]

x can be either a global optimum or a local one, we want to find the global(s) optima.

\vspace{10pt}

To find the solution we will use some optimization algorithm, by definition an optimization algorithm is an iterative method that at the end of each iteration returns a solution \begin{math}
    \in 
\end{math} S. One of the usual algorithm is \textbf{Random search}. 

As already seen we have the no free lunch theorem that stops us from finding a perfect solution. 

\vspace{10pt}
\textbf{No Free Lunch Theorem} \ra 
Let $A_1$ and $A_2$ be two optimization algorithms. 
When performance is averaged uniformly over all possible objective functions, 
$A_1$ and $A_2$ have identical expected performance.

\vspace{10pt}

The consequences of this theorem are that
\begin{enumerate}
    \item It is impossible to have an algorithm that is better than all the others and on all possible problems (no super algorithm can exist).
    \item Given a problem it can't exist an automatic method to find the best algorithm. Then we need to test, experiment and use our prior knowledge.
\end{enumerate}

\vspace{10pt}

Now we will see some algorithms. We start with \textbf{Hill climbing}. Before anything we need to clarify that we can't find every possible solution, so we need to use certain starting points.

\vspace{10pt}

\textbf{Neighborhood.}
Let $N : S \to 2^S$ be a neighborhood function that maps a solution 
$x \in S$ to a set of neighboring solutions $N(x) \subseteq S$.

Starting from an initial solution $x_0$ (typically chosen at random), 
the objective function is evaluated over $N(x_k)$, and the best 
neighbor is selected:
\[
x_{k+1} = \arg\min_{y \in N(x_k)} f(y)
\]
(for a minimization problem).

This process is repeated iteratively until a stopping criterion is met.

\vspace{10pt}

\textbf{Example}

\vspace{10pt}

\begin{itemize}
    \item S = {students in the room}
    \item f = student number 
    \item Maximization problem
    \item Neighborhood = the students around the selected student
\end{itemize}

\textbf{Local optimum} \ra given an optimization problem (S, f) a local optimum is a solution x such that $\forall y \in N(x), f(x)$ is better than $f(y)$. By definition hill climbing will return a local optimum with no guarantee that it will also be global. We can find the global optimum by expanding the radius or by trying more initialization

\vspace{10pt}

In pseudocode (general case)

\begin{enumerate}
    \item Initialize the current solution $x$ (typically at random).
    \item Repeat
    \begin{enumerate}[label=\theenumi.\arabic*]
        \item Generate a neighbor $y$ of $x$.
        \item If $f(y)$ is better than or equal to $f(x)$, set $x := y$.
    \end{enumerate}
    Until for all neighbors $y$ of $x$, $f(y)$ is worse than $f(x)$.
    \item Return $x$.
\end{enumerate}

In strict hill climbing we change the point only if the fitness is strictly better than the previous. With this setting we can have an infinite iteration between 2 equal points. We can set a timeout if we see the same points n times. 

Taboo search can help the search by using memory, avoid the already checked points. 

\vspace{10pt}

\textbf{Example}

\vspace{10pt}

\begin{itemize}
    \item $S = \{ i \in \mathbb{N} : 0 \le i \le 15 \}$
    \item  f(i) number of 1s in the binary code of i
    \item Maximization
    \item Neighborhood = i and j $\iff$ |i - j| = 1
\end{itemize}


We produce a plot

\begin{itemize}
    \item X \ra all solutions produced ordered by their neighborhood
    \item Y \ra all the fitness values
\end{itemize}

\begin{figure}[h!]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=14cm,
    height=6cm,
    xlabel={Solution $i$ (ordered by $|i-j|=1$)},
    ylabel={$f(i)$ = number of 1s in binary$(i)$},
    xtick={0,...,15},
    ymin=0,
    ymax=6,
    grid=major
]

\addplot+[mark=*] coordinates {
(0,0)
(1,1)
(2,1)
(3,2)
(4,1)
(5,2)
(6,2)
(7,3)
(8,1)
(9,2)
(10,2)
(11,3)
(12,2)
(13,3)
(14,3)
(15,4)
};

\end{axis}
\end{tikzpicture}
\caption{Fitness landscape for $S=\{0,\dots,15\}$ with neighborhood $|i-j|=1$.}
\end{figure}

With the fitness landscape we can find the difficulty of the problem. A very smooth one is usually of a simple problem while a very oscillating one will be probably more difficult.


\subsection{Fitness landscape}

In the last class we saw what is a fitness landscape. The plot in the 2-D setting is obtained by plotting on the axis

\begin{itemize}
    \item Horizontal \ra All individual in the search space sorted according to the neighborhood
    \item Vertical \ra the fitness
\end{itemize}

Given an optimization problem (S, f) and a neighborhood N, a solution $x \in S$ is called a local optimum if $\forall y \in N(x)$ f(y) is worse than f(x). By construction hill climbing will return a local optimum with no guarantee that it will also be a global optimum.

\vspace{10pt} 

The fitness landscape gives a visual idea of how easy/hard is for an algorithm to solve the problem
\begin{itemize}
    \item Smooth \ra easy
    \item Rugged \ra hard
\end{itemize}

But in practice we can't draw the landscape because the search space is humongous, and the neighborhood is usually multidimensional.

\vspace{10pt}

Example
\begin{itemize}
    \item $S = \{ i \in \mathbb{N} | 0 \le i \le 15 \}$
    \item f(i) = $i^2$
    \item N = The solutions x and y are neighbors $\iff |x-y| = 1$
    \item Maximization
\end{itemize}

\begin{tikzpicture}
\begin{axis}[domain=0:15]
\addplot {x^2};
\end{axis}
\end{tikzpicture}

Here hill climbing will arrive at the global optimum.

\vspace{10pt}

If we want to look for the number of 1 in the binary code of i we can encounter problems and stay stuck in a local optimum without being able to find the global one. 
\vspace{10pt}

The definition of neighborhood is not fixed, we can define it in 2 ways.

\begin{itemize}
    \item N(x) \ra \{ $y \in s | d(x,y) \le k \}$ \ra distance based
    \item N(x) \ra \{ $y \in s | y = op(x) \}$ \ra action based
\end{itemize}


\begin{center}
\textbf{Example}
\end{center}

\begin{tikzpicture}[scale=1.2]

% axes
\draw[->] (0,0) -- (0,4) node[above] {$f$};
\draw[->] (0,0) -- (10,0) node[right] {$S$};

% deceptive peak near 0
\draw (0.3,3.5) -- (1.5,0);

% small mark around 0
\draw[decorate,decoration={brace,amplitude=5pt}]
(0.1,-0.2) -- (1.6,-0.2) node[midway,below=6pt] {$\approx 0$};

% almost flat line
\draw (1.5,0) -- (6,0.1) -- (8,-0.1) -- (10,-0.3);

% labels
\node at (4,3.5) {Maximization};

% deceptive problem bubble
\draw (7.5,3) ellipse (1.8 and 0.7);
\node at (7.5,3) {Deceptive Problem};

% arrow pointing to plateau
\draw[->] (8.7,2.4) -- (7.8,0.2);

% bottom text
\node at (5,-1) {Random search is ``good'' for this type of problem.};

\end{tikzpicture}

Hill climbing pros

\begin{itemize}
    \item Simple
    \item Efficient
    \item Versatile
\end{itemize}

Cons
\begin{itemize}
    \item Always stops on local optima
\end{itemize}

How to improve hill climbing

\begin{itemize}
    \item Parallel hill climbing
    \item Enlarge the neighborhood
    \item Give the possibility of worsening the fitness of the current solution (within a given probability)
\end{itemize}


\subsection{Annealing}

We now introduce the concept of annealing \ra an experiment aimed at finding the solid state of a material with the minimum level of energy.

\textbf{Metropolis algorithm}

\begin{enumerate}
    \item Consider the initial state of the material with energy $E_i$
    \item Alter the state to get a new energy $E_j$
    \item If $E_j \le E_i$ accept the new state and continue
    \item Else accept the state with probability $exp(-\frac{|E_i - E_j|}{K_B T})$
\end{enumerate}

Where 
\begin{itemize}
    \item T = Temperature
    \item $K_B$ = Boltzmann constant
\end{itemize}

\textbf{Simulated annealing}

\begin{itemize}
    \item Current solution = x
    \item Neighbor = y
\end{itemize}

\[
P(y) = \begin{cases}
    1 \text{ if f(y) better or equal than f(x)} \\
    exp(-\frac{|f(x) - f(y)}{C}) \text{ otherwise}
\end{cases}
\]

Where C is a control parameter

\subsection{Practical applications of Hill climbing}

Test case 1 \ra Given a set of possible investments (N), for each investment we know

\begin{itemize}
    \item Cost
    \item Estimated gain
\end{itemize}

We also know the budget of the client. What subset of the investments should our client do? 

Such that
\begin{itemize}
    \item The total cost is smaller or equal to the budget
    \item The profit is maximized
\end{itemize}


Using a numeric example

\begin{itemize}
    \item N = 10
    \item Costs = \{23, 31, 29, 44, 53, 38, 63, 85, 89, 82 \}
    \item Profits = \{92, 57, 49, 68, 60, 43, 67, 84, 87, 72 \}
    \item Budget = 165
\end{itemize}

\vspace{10pt}
Questions to answer
\begin{enumerate}
    \item What are the solutions ?
    \item How it is convenient to represent the solutions?
    \item What's the fitness?
    \item If the algorithm is neighborhood based, then what is the neighborhood?
\end{enumerate}

Golden rule \ra \textbf{DO NOT solve the problem}

Answers

\begin{enumerate}
    \item All the subsets of the set of investments
    \item As sequences of values, a string of bits of length N. The array will have a pair of  cost and profit. We use 0s and 1s to "1-hot encoding" them. 
    \item The sum for all the indexes where x is 1 of all the profits. $\sum\limits_{i:x[i]=1} p[i]$. We have a maximization problem. We only care about solutions that have a cost less than the budget. The fitness if this condition is not accepted we have -1 or 0 as fitness. Another way to define a non-good fitness is to use the negative sum of the costs such that the distance ourselves as much as possible from a bad solution. 
    \item The neighborhood is obtained here using the bit flip operator. 
\end{enumerate}


\vspace{10pt}

Test case 2 \ra Travel-sale-person: given N cities for which we know all the pairwise distances between themselves. Give one of these cities there is one specific one that is called home. We want the cyclic path that starts from home, visits all the cities once and only once and terminates at home using the shortest possible path.

\vspace{10pt}

We start from a matrix 

\begin{itemize}
    \item N = 5 \ra the matrix is 5x5
    \item Inside the matrix we have the costs/distances
\end{itemize}

Answers 

\begin{enumerate}
    \item Solutions \ra All cyclic paths starting and finishing in home and visiting all cities once. 
    \item Representation \ra An ordered string of the cities
    \item Fitness \ra the sum of the distances in the string
    \item Neighborhood \ra We can use a swapping operator
\end{enumerate}



\subsection{Simulated annealing}

Simulated annealing is similar to hill climbing, but we add the concept of probability. This can lead to a worse fitness. The idea is that we can explore more possible solutions and obtain a better optimum. But we can also miss some of those. Sometimes can be convenient that our current solution is a random neighbor.

\vspace{10pt}

We can start by defining our current solution $x$ and our candidate neighbor $y$.

The probability $P(y)$ is defined as:

\[
P(y) = 
\begin{cases} 
    1 & \text{if } f(y) \leq f(x) \\
    \exp\left(-\frac{|f(x) - f(y)|}{C}\right) & \text{otherwise}
\end{cases}
\]

Where $C$ is a control parameter $C > 0$.

\vspace{10pt}

Afterward, we look for a random number in the interval $[0, 1]$ and if this number is $< P(y)$, we accept the move.

We start with a large $C$, and we decrease it during the algorithm. $C$ will tend to $0$ without reaching it, following the cooling schedule:

\[
C = \frac{C}{h}
\]

Simulated annealing follows the following structure

\begin{enumerate}
    \item Initialize the current solution x (typically at random)
    \item Initialize C and L
    \item Repeat the following until termination
    \begin{itemize}
        \item Repeat L times
        \begin{itemize}
            \item Create a random neighbor x of y
            \item Accept or not y with a certain probability 
        \end{itemize}
    \item Update C (not mandatory update L)
    
    \end{itemize}
    \item Return x (or the best solution so far)
\end{enumerate}

We terminate when we have found a certain condition that can be either a good enough solution or we have run a prefixed maximum number of iterations of the external loop.

\vspace{10pt}

\textbf{Example}

\begin{itemize}
    \item S = {i $\in \mathbb{N} | 0 \le i \le 15$ }
    \item $\forall i \in S, f(i) = $ number of 1s in the binary code of i
    \item Maximization
    \item Neighborhood \ra x and y are neighbors $\iff |x-y|=1$
\end{itemize}

\noindent
\textbf{Step 1: Initialization} \\
$x = 5$ (random initialization) $\rightarrow bin(5) = 101$ \\
$f(x) = 2$ \\
$N(x) = \{4, 6\}$ \\
\textbf{Assume} $y = 6$ (random neighbor) $\rightarrow bin(6) = 110$ \\
$f(y) = 2$ \\
Since $f(y) \leq f(x)$, we accept the move: $x = 6$.

\vspace{10pt}

\noindent
\textbf{Step 2: Iteration} \\
$x = 6$ \\
$N(x) = \{5, 7\}$ \\
\textbf{Assume} $y = 7$ (random neighbor) $\rightarrow bin(7) = 111$ \\
$f(y) = 3$ \\
Since $f(y) > f(x)$, we calculate the acceptance probability.

\vspace{10pt}

\noindent
\textbf{Step 3: Probability Calculation} \\
Assume in this phase the control parameter is $C = 1$. \\
Current state: $f(x) = 3$ \\
Candidate neighbor: $f(y) = 1$

\[
P(acc \ y) = \exp\left( -\frac{|3 - 1|}{1} \right) = e^{-2} \simeq 0.13
\]

\vspace{5pt}
\noindent
This value represents a \textit{``large'' probability} for this phase of the algorithm.

\vspace{10pt}

\textbf{Theorem of asymptotic convergence of simulated annealing}

\vspace{10pt}

Given an optimization problem (S, f) and an appropriate neighborhood, if we decrease C so that it tends to 0 (without reaching it):

\noindent
As $t \rightarrow \infty$ and $C \rightarrow 0$, we analyze the convergence:

\[
\lim_{C \to 0} q_i(C) = \frac{1}{|S_{opt}|} \cdot g(i)
\]

\noindent
\textbf{Where:}
\begin{itemize}
    \item $q_i(C)$ is the probability that Simulated Annealing (SA) will stabilize on solution $i$ with control parameter $C$.
    \item $S_{opt}$ is the set of global optima.
    \item $g(i)$ is an indicator function defined as:
\end{itemize}

\[
g(i) = 
\begin{cases} 
    1 & \text{if } i \in S_{opt} \\
    0 & \text{otherwise}
\end{cases}
\]

\vspace{10pt}
\noindent
This limit shows that as the temperature $C$ approaches zero, the probability of being in a state $i$ becomes uniform over the set of global optimal solutions and zero elsewhere. To be sure to find a global optimum we should execute the process a very large number of times.

\vspace{10pt}

The parameters L and C are respectively Large and slow (decreasing)

\subsection{Genetic algorithms}

Genetic algorithms share similarities with simulated annealing

\begin{itemize}
    \item Both are optimization algorithms
    \item Both allow to look for good solutions without an exhaustive analysis of the search space
    \item Both start from random solutions and improve their quality iteratively
\end{itemize} 

The differences on the other hand are

\begin{itemize}
    \item The solutions (individuals) that are considered at a time are more than 1 (population)
    \item Inspired by the theory of evolution of Charles Darwin
    \item Individuals must be represented by strings of constant length
\end{itemize}


The theory of evolution has the following key elements
\begin{itemize}
    \item Reproduction
    \item Ability of adaptation
    \item Inheritance
    \item Variation
    \item Competition
\end{itemize}

\begin{figure}[h]
    \centering
    \begin{tikzpicture}[
        node distance=2cm and 3cm,
        block/.style={rectangle, draw, minimum width=3cm, minimum height=2cm, align=center, thick},
        arrow/.style={-Stealth, thick}
    ]

    % Nodi (I box del tuo disegno)
    \node[block] (P) {Pop. $P$ \\ of $N$ individuals};
    \node[block, right=of P] (P1) {Pop. $P'$ \\ of $N$ individ.};
    \node[block, below right=1cm and 0.5cm of P] (P2) {Pop. $P''$ \\ of $N$ indiv.};

    % Etichette sopra i box
    \node[above=0.1cm of P1, font=\small] {POP. OF PARENTS};
    \node[below=0.1cm of P2, font=\small] {POP. OF OFFSPRING};

    % Frecce con etichette
    \draw[arrow] (P) -- node[above] {SELECTION} node[below] {PROCESS} (P1);
    
    \draw[arrow] (P1.south) .. controls +(0,-2) and +(2,0) .. 
        node[right, align=left, xshift=0.5cm] {REPRODUCTION \\ WITH VARIATION \\ (CROSSOVER AND \\ MUTATION)} (P2.east);
        
    \draw[arrow] (P2.west) .. controls +(-2,0) and +(0,-2) .. 
        node[left, xshift=-0.5cm] {$P := P''$} (P.south);

    \end{tikzpicture}
    \caption{Genetic Algorithms Cycle}
\end{figure}

\noindent
The \textbf{Selection Process} maps the initial population $P$ to the intermediate population $P'$ (parents):

\vspace{10pt}

\begin{center}
    Population $P$ \hspace{2cm} Population $P'$ \\
    \vspace{5pt}
    \fbox{
        \begin{tabular}{cc}
            A & B \\
            C & D
        \end{tabular}
    }
    \hspace{1.5cm} $\xrightarrow{\text{Selection}}$ \hspace{1.5cm}
    \fbox{
        \begin{tabular}{cc}
            A & A \\
            A & C
        \end{tabular}
    }
\end{center}

\vspace{10pt}

\noindent
In this example, individual \textbf{A} is selected three times due to its high fitness, \textbf{C} is selected once, while \textbf{B} and \textbf{D} are discarded.

\vspace{10pt}

The selection process is made of iterations of N independent executions of a selection algorithm. The selection algorithm chooses 1 solution from a population P.

The algorithm has the following general principles

\begin{enumerate}
    \item It's probabilistic
    \item Given any pair of independent A and B in P if f(A) is better than f(B) then A has a bigger probability than B of being selected
    \item No individual in P has a probability equal to 0 of being chosen
\end{enumerate}

\noindent
\textbf{Fitness Proportionate (Roulette Wheel) Selection} \\
Given $N$ individuals in population $P = \{1, 2, \dots, N\}$, for each $i = 1, 2, \dots, N$ we define the probability of selection as:

\[
P(sel \ i) = \frac{f_i}{\sum_{j \in P} f_j}
\]

\vspace{10pt}

\noindent
\textbf{Mechanism ($N = 4$):} \\
This method acts like a roulette wheel where the area of each slice $i$ is proportional to its fitness $f_i$. 
Individuals with higher fitness occupy a larger portion of the wheel, increasing their chances of being selected when the wheel is "spun" (represented by a random number $U \in [0, 1]$).

\begin{itemize}
    \item $i_1, i_2, i_3, i_4$: Individuals in the population.
    \item $U$: Random pointer used to select the individual.
\end{itemize}

Other ways that can be used are \textbf{Ranking} and \textbf{Tournament} selection.




\subsection{Again Genetic algorithms}

We forgor something about simulated annealing. The theorem that we talked about miss something.

Basically we can only use simulated annealing in an appropriate algorithm, but we didn't define what we meant by appropriate. 

\vspace{10pt}

From the part of asymptotic convergence. We must have a certain property \ra given any pair of solutions x and y it must be possible to build a sequence of solutions $a_1, a_2,..., a_n$ such that $\forall i = 1,2,3,...,n-1, a_i \text{ and } a_{i+1} $are neighbors. So $a_1=x$ and $a_n=y$. We say that the neighborhood is completely connected. 



\vspace{10pt}

Now going back to genetic algorithms.

We start with a population P of N individuals. N is a hyperparameter. We modify the population with two steps.
\begin{itemize}
    \item Selection phase \ra We have a population P' of N individuals
    \item Reproduction with Variation phase \ra We have a population P'' of N individuals
\end{itemize}

The new current population will be P'' and we repeat the two phases until we reach a stopping criterion, and we select as solution the best individual in the final P. 

We create new individuals (offspring) via means of crossover and mutation.

\vspace{10pt}

The idea behind this algorithm comes from the Darwinian ideas of adaptation, competition, reproductions, inheritance and variation. The first two are for the first phase, while the other 3 for the second one. The selection phase is an iteration of N independent executions of a selection algorithm where the objective is to choose one solution from a population of N solutions. 

\vspace{10pt}

Is important that the selection algorithm is probabilistic since we want to have different solutions every time. The probability of selecting a solution must be based on fitness. All individuals must have a non-zero probability of being selected. 

\vspace{10pt}

In this course we will talk about 3 main selection algorithms

\subsubsection{Fitness proportionate selection}

Also known as roulette wheel. We have a population of N individuals called 1, 2,..., n. For each i in 1, 2,..., n we have that the probability of selecting i we have $P(i) = \frac{f(i)}{\sum_{j \in P} f(j)}$. This is a maximization only algorithm, for minimization we can do 1 divided by fitness but if we have 0 fitness we use to add a constant to every possible value. 

\vspace{10pt}

The general take out concept is to adapt our solutions to every case. 


\begin{tikzpicture}
    % Definizione dei colori per i diversi individui
    \def\colorA{orange!60}
    \def\colorB{blue!40}
    \def\colorC{green!50}
    \def\colorD{red!40}

    % Disegno delle fette (angoli calcolati in base alla fitness)
    % Individuo 1 (40% -> 144 gradi)
    \filldraw[fill=\colorA, draw=black] (0,0) -- (0:3) arc (0:144:3) -- cycle;
    % Individuo 2 (25% -> 90 gradi)
    \filldraw[fill=\colorB, draw=black] (0,0) -- (144:3) arc (144:234:3) -- cycle;
    % Individuo 3 (20% -> 72 gradi)
    \filldraw[fill=\colorC, draw=black] (0,0) -- (234:3) arc (234:306:3) -- cycle;
    % Individuo 4 (15% -> 54 gradi)
    \filldraw[fill=\colorD, draw=black] (0,0) -- (306:3) arc (306:360:3) -- cycle;

    % Etichette centrali (Individui e Probabilità)
    \node at (72:2) {\textbf{Ind. 1}};
    \node at (72:1.5) {$P_1 = 0.40$};
    
    \node at (189:2) {\textbf{Ind. 2}};
    \node at (180:1.5) {$P_2 = 0.25$};
    
    \node at (270:2) {\textbf{Ind. 3}};
    \node at (270:1.5) {$P_3 = 0.20$};
    
    \node at (333:2) {\textbf{Ind. 4}};
    \node at (333:1.5) {$P_4 = 0.15$};

    % Disegno della freccia di selezione (Pointer)
    \draw[line width=2pt, -stealth, red] (4,0) -- (3.2,0);
    \node[anchor=west] at (4.1,0) {\textbf{Selection Point}};

    % Titolo o decorazione centrale
    \draw[fill=white, draw=black, thick] (0,0) circle (0.5);
    \node at (0,0) {$f_i$};

\end{tikzpicture}

\subsubsection{Ranking selection}

We follow this procedure

\begin{enumerate}
    \item Sort the individuals from the worst to the best based on fitness. 
    \item We calculate the probability selection based on the indexes. So $P(i) = \frac{\text{index}}{\sum \text{indexes}} $
\end{enumerate}

This algorithm is not sensitive to fitness differences, we don't care about how much a certain individual is better than another one. We can generalize this algorithm by changing the function (now is linear) to have a logarithm, an exponential and so on. This can change the selection pressure. 


\begin{tikzpicture}
    % Definizione colori
    \def\colorRankOne{blue!60}
    \def\colorRankTwo{blue!45}
    \def\colorRankThree{blue!30}
    \def\colorRankFour{blue!15}

    % Rappresentazione a "gradini" o Ruota basata su Rank
    % Supponiamo 4 individui. Somma ranghi = 1+2+3+4 = 10.
    % Spazi: 4/10 (144°), 3/10 (108°), 2/10 (72°), 1/10 (36°)

    % Rank 1 (Il migliore) - 144 gradi
    \filldraw[fill=\colorRankOne, draw=black] (0,0) -- (0:3) arc (0:144:3) -- cycle;
    % Rank 2 - 108 gradi
    \filldraw[fill=\colorRankTwo, draw=black] (0,0) -- (144:3) arc (144:252:3) -- cycle;
    % Rank 3 - 72 gradi
    \filldraw[fill=\colorRankThree, draw=black] (0,0) -- (252:3) arc (252:324:3) -- cycle;
    % Rank 4 (Il peggiore) - 36 gradi
    \filldraw[fill=\colorRankFour, draw=black] (0,0) -- (324:3) arc (324:360:3) -- cycle;

    % Etichette
    \node at (72:2.2) {\textbf{Rank 1}};
    \node at (72:1.7) {$P = 0.4$};
    
    \node at (198:2.2) {\textbf{Rank 2}};
    \node at (180:1.7) {$P = 0.3$};
    
    \node at (288:2.2) {\textbf{Rank 3}};
    \node at (288:1.7) {$P = 0.2$};
    
    \node at (342:2.2) {\textbf{Rank 4}};
    \node at (350:1.7) {$P = 0.1$};

    % Pointer
    \draw[line width=2pt, -stealth, gray] (4,0) -- (3.2,0);
    \node[anchor=west] at (4.1,0) {Selection};

    % Centro
    \draw[fill=white] (0,0) circle (0.4);
    \node at (0,0) {$\mathcal{R}$};

\end{tikzpicture}

\subsubsection{Tournament selection}

We follow the following schema (To draw). 


This can be considered as both random and deterministic.

\begin{enumerate}
    \item It is possible (in some cases) to not evaluate all the possible individual's fitness.  
    \item We can change the selection pressure by changing the tournament size
\end{enumerate}


\begin{tikzpicture}[
    node distance=1.5cm,
    individual/.style={rectangle, draw, fill=gray!10, minimum width=2cm, minimum height=0.8cm, rounded corners},
    winner/.style={rectangle, draw=green!60!black, fill=green!10, minimum width=2cm, minimum height=0.8cm, rounded corners, line width=1.5pt},
    loser/.style={rectangle, draw=red!40, fill=red!5, minimum width=2cm, minimum height=0.8cm, rounded corners, dashed}
]

    % Titolo del Torneo
    \node[font=\Large\bfseries] at (2.5, 2) {Tournament Selection ($k=3$)};

    % Popolazione di sfondo (simbolica)
    \draw[dashed, gray] (-1,-4.5) rectangle (1.5, 0.5);
    \node[gray, rotate=90] at (-0.7,-2) {Population Pool};

    % Individui estratti per il torneo
    \node[individual] (ind1) at (3, 0) {Ind. A (Fit: 45)};
    \node[individual] (ind2) at (3, -1.5) {Ind. B (Fit: 82)};
    \node[individual] (ind3) at (3, -3) {Ind. C (Fit: 12)};

    % Frecce di estrazione dalla pool
    \draw[latex-, shorten <=2pt] (ind1.west) -- (1.5, 0);
    \draw[latex-, shorten <=2pt] (ind2.west) -- (1.5, -1.5);
    \draw[latex-, shorten <=2pt] (ind3.west) -- (1.5, -3);

    % Nodo del Vincitore
    \node[winner] (winner) at (10, -1.5) {Winner: Ind. B};

    % Frecce del confronto
    \draw[-Stealth, thick] (ind1.east) -- (winner.west) node[midway, above, sloped, font=\small] {loss};
    \draw[-Stealth, ultra thick, green!60!black] (ind2.east) -- (winner.west) node[midway, above, font=\small] {Best Fitness};
    \draw[-Stealth, thick] (ind3.east) -- (winner.west) node[midway, below, sloped, font=\small] {loss};

    % Nota sulla Pressione Selettiva
    \node[draw, fill=yellow!10, text width=4cm, font=\footnotesize, anchor=north] at (5, -4.5) {
        \textbf{Logic:} \\
        Pick $k$ individuals at random. \\
        The one with the highest fitness wins the tournament.
    };

\end{tikzpicture}


\subsection{Crossover and mutation}

Now we talk about the part about reproduction with variation. Selection depends only on phenotype while crossover and mutation only on genotype (DNA structure).


\subsubsection{Crossover}

We start by taking two solutions that are in P'. 

\begin{tikzpicture}
    \tikzset{bit/.style={rectangle, draw=black, minimum size=0.9cm, font=\ttfamily\bfseries}}
    \tikzset{highlight1/.style={fill=blue!20}}
    \tikzset{highlight2/.style={fill=orange!20}}

    % ── PARENTS ──────────────────────────────────────────
    \node[anchor=west, font=\bfseries] at (0, 2.2) {Parents};

    % Etichette riga
    \node[anchor=east] at (0.8, 1.3) {P1:};
    \node[anchor=east] at (0.8, 0.2) {P2:};

    % P1
    \foreach \val [count=\i] in {1,0,1,1,0,0,1} {
        \node[bit, highlight1] at (\i*0.9+0.8, 1.3) {\val};
    }

    % P2
    \foreach \val [count=\i] in {0,1,0,0,1,1,0} {
        \node[bit, highlight2] at (\i*0.9+0.8, 0.2) {\val};
    }

    % ── LINEA DI TAGLIO (tra 3° e 4° bit) ───────────────
    % Centro del 3° bit: x = 3*0.9+0.8 = 3.5
    % Centro del 4° bit: x = 4*0.9+0.8 = 4.4
    % Punto di taglio: (3.5+4.4)/2 = 3.95
    \draw[red, ultra thick, dashed] (3.95, 2.8) -- (3.95, -3.5);
    \node[below, font=\small\sffamily, text=red] at (3.95, -4.0) {Cutting point};

    % ── OFFSPRING ────────────────────────────────────────
    \node[anchor=west, font=\bfseries] at (0, -1.2) {Offspring};

    % Etichette riga
    \node[anchor=east] at (0.8, -1.8) {F1:};
    \node[anchor=east] at (0.8, -2.9) {F2:};

    % F1: primi 3 da P1 + ultimi 4 da P2
    \foreach \val [count=\i] in {1,0,1} {
        \node[bit, highlight1] at (\i*0.9+0.8, -1.8) {\val};
    }
    \foreach \val [count=\i] in {0,1,1,0} {
        \node[bit, highlight2] at (\i*0.9+3.5, -1.8) {\val};
    }

    % F2: primi 3 da P2 + ultimi 4 da P1
    \foreach \val [count=\i] in {0,1,0} {
        \node[bit, highlight2] at (\i*0.9+0.8, -2.9) {\val};
    }
    \foreach \val [count=\i] in {1,0,0,1} {
        \node[bit, highlight1] at (\i*0.9+3.5, -2.9) {\val};
    }

\end{tikzpicture}


This example is for 1-point crossover, but we can have multiple cutting points. 


\subsubsection{Mutation}

For each character we change it with a given probability of mutation (another hyperparameter)


\subsubsection{General function of genetic algorithms}

Here we have a pseudocode on the general idea of genetic algorithms

\begin{enumerate}
    \item Initialize a population P of N individuals (typically at random)
    

    \item Repeat until termination condition is reached:
    

    \begin{enumerate}
        \item Calculate the fitness of all the individuals in the population P (can be avoided in some cases, like tournament selection). This step is computationally costly
        \item We create an empty population P'
        \item Repeat P' has N individuals
        \begin{enumerate}
            \item We choose a genetic operator (crossover with probability $P_c$ or replication with probability $1-P_c$)
            \item Select 2 individuals from population P 
            \item Apply the operator to the individuals 
            \item Apply mutation to the individuals resulting from the previous step 
            \item Insert the individuals resulting from the previous point into P' 
        \end{enumerate}
        \item Replace P with P'
    \end{enumerate}
    \item Return the best individual in P in terms of fitness
\end{enumerate}



Some points

\begin{itemize}
    \item If N is odd:
    \begin{itemize}
        \item Select 1 individual mutated and insert it into P'
        \item One crossover event copies into P' only one offspring (typically the best in fitness)
    \end{itemize}
    \item Elitism is the unchanged copy of the best K individuals in P into P' (bypassing the selection-crossover-mutation pipeline), typically K=1
    \item We have to do something like P \ra selection \ra P' \ra crossover + mutation \ra P'' \ra P := P'' 
    \item Replication is the unchanged copy of the parents
    \item Termination condition is either
    \begin{itemize}
        \item A good enough solution is found, and we return it
        \item A maximum number of generation was performed
    \end{itemize}
\end{itemize}


\subsubsection{Hyperparameters of genetic algorithms}

\begin{itemize}
    \item String length \ra Given by the problem definition
    \item Population size (N) \ra The bigger, the better. We have an upper bound (true for every problem)
    \item Selection algorithm that we want to use \ra No free lunch holds, but tournament selection can be interesting to use (fast, and we can use the selection pressure). We can start with this and try the others. The tournament size should not be too big (2 can be good, and the minimum). 
    \item Crossover probability ($P_c$) \ra Should be close to 1
    \item Mutation probability ($P_m$) \ra Should be close to 0
    \item Maximum number of generations \ra The bigger, the better but with an upper bound like for N
    \item Elitism on or off  (and if on size of the elite and k) \ra Elitism should be used with K=1
\end{itemize}


