\section{Database}

\subsection{Theory}

\subsubsection{Introduction}

A database can be defined as a collection of related data items within a specific business.
DBMS are sotwares that define, create, use and maintain a database, if we combine a DBMS with a database we get a database system
The majority of DBMS softwares rely on SQL systems. 

\vspace{10pt}

To work with databases we need to model them to understand how they work. Modelling means representing reality in a 
scaled version. The same with databases we connect information to represent how relationships in the world exist.
Based on the business needings we design the database in a logical fashion and we translate it to 
a our language of choice. The step is passing from conceptual design to logical design. The first is DBMS independent while the second is DBMS dependent.
The process is:

\begin{enumerate}
    \item Requirement collection and analysis
    \item Conceptual design
    \item Logical design
    \item Physical design
\end{enumerate}

\vspace{10pt}

An SQL database is based on entities and relationships. Entities are business concepts
with an unambiguous meaning to a particular set of users. Entities have certain characteristics called attributes.
The rows are the entities (also called instances). The columns are the attributes. An association 
between one or more entities is called relationship, it defines how tables are related to each other. 

\vspace{10pt}

Another useful concept is the one of cardinality, how many instances of the entities can be associated with each other. 

We can have
 \begin{itemize}
    \item 1 to 1
    \item 1 to many
    \item Many to many
\end{itemize}

\vspace{10pt}

An important step in the design is the one related to data normalization, we want to avoid redundant data. We split 
the tables into smaller tables to avoid anomalies during insertion, deletion or updates. The 
procedure follows a step by step transformation. The database is more complex from the structural point of view but can be easier to manage.

We have 3 possible normal forms.

\begin{itemize}
    \item First NF \ra Each cell in the table can have only one value, never a list of values 
    \item Second NF \ra An entity type is in 2NF when it is in 1NF and when all of its non-key
attributes are functionally dependent on its primary key. Features should be fully dependent on the entire primary key
    \item Third NF \ra An entity type is in 3NF when it is in 2NF, and non-key attributes are
directly (non-transitively) dependent on the primary key.
In other words: non-prime attributes must be functionally dependent on the
key(s), but they must not depend on another non-prime attribute.

\end{itemize}

