\section{Data Mining}


\subsection{Introduction}

Fernando Becao presents himself and the tutor.

Fernando Becao asks us why we weren't at the party.

We study about supervised and unsupervised learning.

Exam + project.

Exam is 55\% of the final grade

Project is 45\% and is divided into

\begin{itemize}

\item Deliverable 1 -> 30\% by November 4th
\item Deliverable 2 -> 60\% by January 3rd
\item Discussion -> 10\%
\end{itemize}

Both exam and project will expire after the year


\subsection{Context}

Random debate about things

We need to find relevant data to reduce computation and improve the quality of works

\vspace{10pt}

\textbf{Big Data} -> we simplify datasets that are too big for normal processing. From big data into mean datasets

Big data characteristics:
\begin{itemize}
    \item \textbf{Volume:} big data are big
    \item \textbf{Velocity:} Big data technology allows databases to process and analyze data while it is being generated
    \item \textbf{Variety:} Big data can be a mixture of structured, unstructured and semi-structured data. To solve this problem Big Data is flexible
\end{itemize}

Some information about AI, ML and Data Science.

\vspace{10pt}


\textbf{AI:} Making things that show human intelligence. Automatize human tasks.

\vspace{10pt}

\textbf{Machine Learning:} Approach AI using systems that can find patterns from data and examples. ML systems learn by themselves. We can see them as a sort of subset of AI or a way towards AI 

\vspace{10pt}

\textbf{Data Science:} the study of where information comes from and how to turn it into a valuable resource.
\vspace{10pt}


\textbf{Data Science vs Data Mining}
\vspace{10pt}

Data science is a set of principle that guide the extraction of information from data. We try to view problems from a data perspective.

\vspace{10pt}

Data mining on the other hand is the extraction of knowledge from data via the use of algorithms.


\vspace{10pt}


Find/build attributes is very important. Basically the properties of the things that we are studying have to be selected or preprocessed.

\vspace{10pt}

After the construction of the features a ML model can learn how to divide the instances on an hyperplane. This can be used to make predictions. Recall that a model will predict only based on the class that it was trained on.

\vspace{10pt}

Is very important to use the relevant and appropriate features. More features can mean more possibility of discriminating between the classes. 

\vspace{10pt} Typically we use labeled examples, enough data and clear cut definitions.

\vspace{10pt}

If our models try to attribute a label we have supervised learning. Future examples will get a prediction.


\vspace{10pt}


To build features we use data warehouse and ETL (extract, transform and load). Recall that we can transform every relational schema into a table (at least with SQL).

\vspace{10 pt}

Minimum information to identify a costumer:
\begin{itemize}
    \item Transaction number
    \item Date and time of transaction
    \item Item purchased
    \item Price
    \item Quantity purchased
    \item A table that matches product code to its name, subgroup code to name, product group code to group name.
    \item Product taxonomy to link product code to subgroup code and product subgroup code to product group code.
    \item Card ID
\end{itemize}

Consistent behaviors are easier to analyze. To found the level of consistency i can use the standard deviation and I can remove the outliers.

\vspace{10pt}

We can also look at relevant variables like:
\begin{itemize}
    \item Recency
    \item Frequency
    \item Monetary value
    \item Average purchase
    \item Most frequent store
    \item Average time between transactions
    \item Standard deviation of transactional interval
    \item Costumer stability index
    \item Relative spend on each product
\end{itemize}

\subsection{Canonical tasks in data mining}

If we want to classify new data from a decision criterion previously learned we talk about \textbf{Supervised learning}

\vspace{10pt}

If we want to summarize a data set we talk about \textbf{Unsupervised learning}

\vspace{10pt}

\textbf{Supervised learning:}
\begin{itemize}
    \item Classification
    \item Regression
\end{itemize}

\textbf{Unsupervised learning:}
\begin{itemize}
    \item Clustering
    \item Visualization
    \item Association
\end{itemize}

\vspace{20pt}

In our datasets the features are the columns while the rows are the instances

\vspace{20pt}

\textbf{Clustering} -> We plot the instances on a hyperplane and we try to group them by distance, we can have different plots 

\textbf{Association rules} -> based on our data we can use the transactions to find some rule to infer costumer routine. We use confidence, support, lift and so on.

\textbf{Visualization} -> n-D data can be difficult to visualize so we can flatten them into 2-D ones. Examples are histograms, bubbles and goggle boxes. We can also do some dimensionality reduction using PCA or other algorithm.

\subsection{The data mining process}


We can have different methodologies to acquire insights from data.

\vspace{10pt}

\textbf{KDD}
\begin{enumerate}
    \item Data
    \item Selection
    \item Preprocessing
    \item Transformation
    \item Data mining
    \item Interpretation/Evaluation
    \item Knowledge  
\end{enumerate}

\textbf{CRISP-DM}
\begin{enumerate}
    \item Business understanding
    \item Data understanding
    \item Data preparation
    \item Modeling
    \item Evaluation
    \item Deployment
\end{enumerate}



\vspace{10pt}


We use ML algorithms instead of fixed formulas because the events are too complex for a simple formula.

We don't understand the problems but we try to approximate solution. Black Box ML approach basically.

\vspace{10pt}

If the model is not good enough we can either improve the model or gather more data. The second one is quicker. Dumb algorithms with a lot of data will learn better than smart ones with not a lot of them.

\vspace{10pt}

Traditional statistics might be described as being characterized
by data sets which are small and clean, which are static,
which were sampled in an iid manner, which were often
collected to answer the particular problem being addressed,
and which are solely numeric.

\begin{table}[h!]
\centering
\scriptsize % riduce la dimensione del font
\renewcommand{\arraystretch}{1.1} % meno spazio tra le righe
\begin{tabular}{p{2.5cm} p{5cm} p{5cm}}
\hline
\textbf{Dimension} & \textbf{Primary data} & \textbf{Secondary data} \\
\hline
Definition & Data you collect yourself for a specific, current purpose. & Data collected by others for a different (often past) purpose. \\
Typical sources & Surveys, experiments, interviews, field measurements, sensors. & Government statistics, research papers, data portals, company databases, syndicated datasets, web-scraped corpora. \\
Control over design & Full control (sampling, instruments, definitions, timing). & Little/no control; must accept others’ design choices. \\
Fit to your question & High: tailored to your problem and target population. & Varies: often indirect or requires redefinition/derivations. \\
Cost & Usually higher (money, time, staff, tooling). & Usually lower or free; licensing may apply. \\
Time to obtain & Longer (planning → collection → cleaning). & Faster (download/access + cleaning/understanding). \\
Timeliness/recency & Up-to-date by design. & May be outdated; release lags common. \\
Granularity & Exactly what you need (variables, frequency, detail). & Fixed by source; may lack key variables or be too aggregated. \\
\hline
\end{tabular}
\caption{Comparison between primary and secondary data.}
\end{table}


\vspace{20pt}

Size of data set is also useful. If we have big datasets even tiny effects exists. They can be useless. Now we should ask if the effect is important or not. 

\vspace{10pt}

From statistical significance to substantive significance.

\vspace{10pt}

Since the datasets are very big we try to compute them in a adaptive or sequential way. Also we might have multiple and interrelated files

\vspace{10pt}


We have two main ways to process data

\begin{itemize}
    \item Incremental
    \item Batch
\end{itemize}

Other characteristics of problems in data mining are

\begin{itemize}
    \item Nonstationarity and population drift
    \item Selection bias
    \item Spurious relationships
\end{itemize}


\vspace{10pt}

Input variables should be causally related to the output. If we have a small number of observations we will have high correlation.

Confounding variables will correlate both with dependent and independent variables. The confounding factor will estimate incorrectly the relation. A \textbf{Spurious relationship} happens when we perceive a relationship between two variables that actually doesn't exist. We ar not accounting for the confounding factor. 

\vspace{10pt}

Input variables must be causally related to the output to be meaningful.

\vspace{10pt}

We have to discriminate between \textbf{causality} and \textbf{correlation}:
\begin{itemize}
    \item Correlation -> two things co-occurs, changing one of them will not change the output
    \item Causality -> a change on the input will cause a change on the output
\end{itemize}

Correlation is a pattern, causation a consequence


\vspace{10pt}

\textbf{Input Space} -> is the input feature vector, the algorithm will look for a solution

\vspace{10pt}

\textbf{Curse of dimensionality} -> bad effects can be caused by redundant features, bigger dimensionality means bigger and more sparse input. Clustering can be harder. Generalization is exponentially harder. Feature selection can be an answer.

\vspace{10pt}

\textbf{Input space coverage} -> representative training examples will improve our model quality. Test data outside the training input space will be bad for the performance.

\vspace{10pt}

\textbf{Interpolation} -> predictions in the range of data that we have

\textbf{Extrapolation} predictions outside the range of data that we have, Time series

\textbf{Separation} -> if we plot our data on a 2-D space we can be able to draw by hand the regions where the data are. ML models will try to do it mathematically. If we can use a linear hyperplane we have \textbf{Linearly separable data} if not they are not linearly separable. With separable classes we can get 0 errors. We want to minimize the error (finding the Bayes error). Simple algorithm like perceptron will solve problems with separables classes.

\vspace{10pt}

Kind of variables
\begin{itemize}
    \item Nominal
    \item Ordinal
    \item Discrete
    \item Continuous
    \item Interval
    \item Ratio
\end{itemize}

\vspace{10pt}

\textbf{Metadata} -> Information that provides information about data

\begin{itemize}
    \item Descriptive metadata
    \item Structural metadata
    \item Administrative metadata
\end{itemize}

\subsection{Exploration}

Today visualization
\vspace{10pt}
Strong points of good visualization

\begin{itemize}
    \item Can show processes
    \item Show comparisons between numbers
    \item Can show differences and changes
    \item Can use geography to show local data
    \item Can show relationships
    \item Can show density
\end{itemize}

\vspace{10pt}

What not to do

\begin{itemize}
    \item Don't clog the graph
    \item No 3-D
    \item Don't use unfaithful charts
\end{itemize}

\vspace{10pt}

Guidelines

\begin{itemize}
    \item Reduce chartjunk
    \item Increase data-ink ration
    \item Use similar structures in the same charts
\end{itemize}

\vspace{10pt}

\textbf{Tufte lie factor} -> Measure of distortion in a graph

\begin{equation}
    \text{Lie Factor} = 
    \frac{\text{Size of effect in graph}}
         {\text{Size of effect in data}}
\end{equation}

\vspace{10pt}

As rule of thumb we should have 0.95 < Lie Factor < 1.05

\textbf{Suggestions}

\begin{itemize}
    \item White background
    \item Don't use colors if not necessary, Charts should be like man suits (Prof, 2025)
    \item Order the items in a smart way
    \item Use a scale
    \item Crisp borders
    \item No smooth line
    \item Simple is better
    \item No 3-D
    \item Discriminate clearly the clusters with shapes and colors
    \item Spaghetti chart should highlight interesting patterns
    \item Clutterplot should highlight interesting points
    \item Use shadows to highlights interesting points
\end{itemize}

\vspace{10pt}

\textbf{Charts that can be useful}

\begin{itemize}
    \item Bar charts, vertical and horizontal
    \item Line charts
    \item Mix of the previous
    \item Stacked bar charts
    \item Scatterplot, also with tendency lines, grids, bubbles
    \item Pie charts, can mix them with histograms, we can label them, compare with others (same as paired column chart). They are like stacked bar charts. We also have part to whole mini pie charts. We can plot them on a graph.
    \item  Radar plot, easy to compare more of them
\end{itemize}

\vspace{10pt}

\textbf{Visualization for analysis}


\begin{itemize}
    \item Histogram, can be stacked and combined with scatterplots
    \item Boxplot, can be combined with scatterplots, histograms
\end{itemize}

\vspace{10pt}

\textbf{Correlation matrices}
\begin{itemize}
    \item Can be done with distributions, scatter plots and matrices
\end{itemize}

\textbf{Parallel coordinate}
\begin{itemize}
    \item Show patterns that can be compared easily
\end{itemize}


\vspace{10pt}

\textbf{Small Multiples}

\begin{itemize}

\item Can show many graphs together, easy for comparisons
   
\item Basically we can have a lot of variables in a 2-D graph

\end{itemize}

\vspace{10pt}

\textbf{Heat maps}
\begin{itemize}
    \item Can show quantities in the plane that we are visualizing
\end{itemize}

\vspace{10pt}

\textbf{Tree maps}
\begin{itemize}
    \item We can see quantities with sizes and dividing them points based on certain characteristics
\end{itemize}

\vspace{10pt}

\textbf{Geo-visualization}

\begin{itemize}
    \item We can aggregate data from different geographical point of view, we can use normal maps, bubbles and plot quantities of them. Cartograms are useful for quantities on and densities.
\end{itemize}

\vspace{10pt}

\textbf{Linked Views}

\begin{itemize}
    \item We can put different vies of the dataset together. We can select different subsets to compare them
\end{itemize}

\subsection{Preprocessing}

Today data preparation and pre-processing

\vspace{10pt}

With data pre-processing and preparation we want to make our data useful for the model. All the datasets are made of signal and noise. We want to maximize the signal and reduce as much as possible the noise.

Real data usually are:
\begin{itemize}
    \item Incomplete
    \item Noisy
    \item Inconsistent
\end{itemize}

\vspace{10pt}


\textbf{Preparation}:
\begin{enumerate}
    \item Missing values
    \item Outliers
    \item Discretization and encoding
    \item Imbalanced datasets
\end{enumerate}

\textbf{Pre processing}

\begin{enumerate}
    \item Feature selection \ra Relevance analysis and redundancy removale
    \item Feature engineering
    \item Feature scaling and normalization
\end{enumerate}


\vspace{10pt}

\textbf{Treating missing data}

\vspace{10pt}

Missing values are values that are not available, very common. 

How to deal:
\begin{itemize}
    \item Delete records with missing data, biased
    \item Delete columns with too many missing data, loose information
    \item Manually insert them
    \item Imputing with central tendency metrics, for general and for subsets
    \item Derive them in an objective way
    \item Use a predictive model, like KNN
    \item Use similarity measures
\end{itemize}

\vspace{10pt}

Usually our go to is to use the quickest and simplest option, after we analyze the performance of the model.
If the error is significantly higher on the subset with imputed values than on the one with non imputed we look for other options.



Today treatment of missing data. Again.

\vspace{10pt}

\textbf{Outliers} \ra Data-points that is distant from the other observations.
It can be from variability, experimental errors or extreme cases based on different causes. 
They can come from
\begin{enumerate}
    \item Unusual but correct situations
    \item Incorrecto measurements
    \item Errors in data collection
    \item Lack of code for missing data
\end{enumerate}

Outliers can have an effect on our training, \textbf{Leverage effect}.



To deal with outliers:

\begin{itemize}
    \item Detect them
    \begin{enumerate}
        \item Automatic limitation using a threshold
        \item If we have a standard distribution we do $3\sigma$ + and - Average
        \item Boxplots
        \item Scatterplots for 2-D cases
        \item Multi-D outliers \ra KNN distance, isolation forest, cluster methods, self.organizing maps, dimensionality reduction 
    \end{enumerate}

    \item Treatment
    \begin{enumerate}
        \item Capping/Winsoring
        \item Transformation methods \ra Log distribution
    \end{enumerate}

\end{itemize}


\textbf{Discretization} \ra We pass from continuous values to discrete ones. 

\vspace{10pt}

Basically we bin our data, we can do supervised and unsupervised discretization.

\begin{itemize}
    \item Equal-width binning, unsupervised
    \item Equal-depth binning, unsupervised
\end{itemize}

Choose the number of bins

\begin{itemize}
    \item Square root rule
    \item Rice rule
    \item Business common sense 
\end{itemize}

For supervised discretization
\begin{itemize}
    \item Entropy
\end{itemize}

\textbf{Encoding} \ra To be able to work with categorical variables we need to transform them into numerical ones.

\begin{itemize}
    \item 1-Hot encoding
    \item Binary encoding
    \item Ordinal encoding
\end{itemize}

\textbf{Imbalanced Learning} \ra if our dataset is imbalanced, we have much more instances for a certain class than for another.

\vspace{10pt}

In an imbalanced dataset usually we have \textbf{majority} and \textbf{minority} classes. We use the \textbf{IR} imbalance ratio to consider it. 
We can't use standard learning ethods, they will introduce a bias in favor of the majority class during training (the minority class will contribute less to the maximisation of the objective function).
Accuracy in this setting becomes almost useless. The problems arise when most algorithms assume a balanced class distribution and uniformity of misclassification costs.

\vspace{10pt}

The usual solutions are

\begin{itemize}
    \item Undersampling
    \item Oversampling
    \item Hybrid approaches
\end{itemize}

One of the most used solution is \textbf{SMOTE} (Synthetic Minority Oversampling TEchnique). 

\begin{enumerate}
    \item Randomly selecting a minority class instance
    \item Define the set of k-nearest neighbors
    \item Randomly select another minority class sample in that KNN set
    \item Generate new samples using linear Interpolation
\end{enumerate}

It's important to remember that we should always use real data instead of synthetic ones. The generated ones should be as realsitica as possible and they will always introduce some noise.

\vspace{10pt}

Now we start with \textbf{Data Preprocessing}, we want to reduce the noise and increase the signal.
The idea is to reduce the size of the dataset in terms of the number of features. 
We want to beat the curse of dimensionality since in high dimensions distance metrics will lose meaning.

\vspace{10pt}

To decide what to keep and what to remove we check two principles

\begin{itemize}
    \item Redundancy \ra features that give us the same information about the dataset
    \item Irrelevance \ra do not affect the target
\end{itemize}

To reduce the input space we can
\begin{itemize}
    \item Feature engineering \ra Derive new features from the existent ones
    \item Heuristic feature selection
    \item Correlation matrices
    \item PCA
\end{itemize}

\vspace{10pt}

Since we care about distances we should also aim at having standardized data.
The main ways are

\begin{itemize}
    \item Normalization
    \item Standardization
\end{itemize}


\subsection{Segmentation}


Cluster analysis is basically classify different objects in a finite and smaller subspace, basically binning with extra-steps.
We want to use their differences and similarities to group or separate them.
The objective is to minimize the intraclusters distances while maximising the distances between clusters.

\vspace{10pt}

The difference from classification is that we do not know to which cluster each example is part of, we want to derive those clusters and analyze them.

In clustering we will have a solution, we define a fixed number of clusters. It's different from A Priori grouping since in the latter we define only the rules.

\vspace{10pt}

An example of A Priori grouping is Cohort analysis \ra group observations based on certain characteristics. 

\vspace{10pt}

\textbf{Cell based segmentation} is based on the quantiles and we try to segment our observations on those. 
We can do that using just one variable using the \textbf{Migration matrix}.

\vspace{10pt}

To analyze our data we can use different approaches. One of them is the \textbf{RFM}.
\begin{itemize}
    \item Recency \ra Customers that have purchased more recently will purchase Again
    \item Frequebcy \ra Customers that have purchased more will buy again
    \item Monetary \ra Customers that have spent more will spend again
\end{itemize}

We can use RFM for customers files that contain purchase history. We have two methods

\begin{enumerate}
    \item Exact Quintiles
    \begin{itemize}
        \item Sort the database according to recency and divide into 5 quintiles
        \item Do the same for frequency and monetary
        \item We have 125 cells of equal size
    \end{itemize}
    \item Hard coding
    \begin{itemize}
        \item We divide the categories into exact values
        \item More expensive to program 
        \item We have many quantities from cell to cell
    \end{itemize}
\end{enumerate}

The popularity of RFM comes from its semplicity and capacity to segment customers.

\vspace{10pt}

On the more general case of cluster analysis we need to understand which are the stages of it

\begin{enumerate}
    \item \textbf{Decide the set of variable that we want to use} \ra 
To decide which variable we want to use we have to kniw which is the objective of the segmentation. Since
the type of problem will determine the variables that we will use. If we want to group objects we want to maximise discrimination ability.
The quality of the cluster analysis is stricly based on the variables used. The choice should replicate a theoretical context and use variables that we know will be good discriminators.

    \item \textbf{Decide the similarity or dissimilarity criterion} \ra The usual choice is calculating the difference between the objects using metrics based on euclidean spaces. The differences are based on how much instances differe on numerical values. A measure of distance should obey three axioms: simmetry, distances are always positive and that we have the triangular inequality. The usual choice is the euclidean distance, but we also have the Minkowski, the weighted euclidean and correlation.
    
    \item \textbf{Decide our clustering algorithm}
    
    \item \textbf{Do profilization with the analysis and validation of the resulting solution}
\end{enumerate}

As said before we have to firstly differentiate between A priori grouping and Clustering.

\vspace{10pt}

To cluster our data we can have different approaches, based on different algorithms and ideas. 
\begin{itemize}
    \item Hierarchical methods
    \item Partition methods
    \item DBScan
    \item Shift means
    \item Self-organizing maps
\end{itemize}

\subsection{Clustering techniques}

\subsubsection{Hierarchical Clustering}

With \textbf{Hierarchical clustering} we start from a feature matrix and we calculate a dissimilarity matrix to see the differences. This last matrix is based on the distances between the points.
To aggregate we can use different aggregation rules like single linkage or complete linkage (connect the 2 closest of farthest points) or average linkage or group linkage.
We can obtain also the dendrogram of the data and visualize differences with it.
Hierarchical clustering can have many disadvantages

\begin{itemize}
    \item We can't go back after an operation
    \item Even though is not computational costly we can't correct wrong decisions
    \item Doesn't works really well with large datasets
\end{itemize}

To improve the performance of Hierarchical methods we can analize the links after each partition (CURE or chamaleon) or integrate iterative optimization (BIRCH).

\vspace{10pt}

\subsubsection{Partition methods}

With \textbf{Partition methods} (like k-means algorithm) we try to divide the data into k (decided at the start) different clusters. We will find for each cluster a centroid.
Each group must contain at least one point and each point must belong to exactly one cluster. The algorithm works like this

\begin{enumerate}
    \item We create a random partition based on k
    \item We relocate iteratively moving objects between groups
    \item We decide if the partitioning is good based on the similarities between groups
\end{enumerate}

More specifically

\begin{itemize}
    \item Choose the seed
    \item Associate each individual to a seed
    \item Calculate the centroids
    \item Go back to 2
    \item End when we don't move anymore the centroids
\end{itemize}

With k-means we want to minimize intra-group variance (the SSE, sum of squared errors)
\begin{equation}
\text{SSE} = \sum_{i=1}^{k} \sum_{x \in C_i} \| x - \mu_i \|^2
\end{equation}

We can reduce the SSE by increasing k but good clustering with small k can have a lower SSE than a poor one with higher k.

\vspace{10pt} 

K-means is simple, efficient and linear but is sensitive to outliers, the initial random position and not good enough for non spherical clusters.
One improvement to the problem if initialization is to use K-means++ that is made via a 4 step initialization process.
\begin{enumerate}
    \item Choose the first centroid randomly
    \item Calculate the distance from the centroids for each point
    \item Choose the next centroid probabilistically
    \item Repeat untill we get k centroids
\end{enumerate}

This way can find better centroids and converges faster.

\vspace{10pt}


\subsubsection{DBScan}

Density Based Clustering is a clustering algorithm meant to find outliers. 
The previously seen methods are based on several assumptions that if they are not true will give us suboptimal results. Here with DBSCAN we don't assume any specific shape for our clusters, basically we want to find regions with an high desity of points. We use mainly two parameters:

\begin{enumerate}
    \item $\epsilon$ \ra radius around a point
    \item MinPts \ra minimum number of points required to form a dense region
\end{enumerate}

Steps 
\begin{enumerate}
    \item Pick a point
    \item Check the neighborhood to find core and border points
    \item Expand the cluster to find if the neighbors of a core point are also core point. We apply this step to all found core points.
    \item Stop when we can't expand our cluster
\end{enumerate}


The choice of the hyperparameters is based most on case to case

\vspace{10pt}

A different approach is HDBSCAN where we can try different values. It adapts at different levels of density both wit $\epsilon$ and MinPts. Basically we want to explore with different density levels and we build a dendrogram.
The parameters are min\_cluster\_size and min\_samples. The steps are

\begin{itemize}
    \item Transform distances
    \item Build a graph and hierarchy
    \item Condense and select clusters
    \item Output
\end{itemize}

\subsubsection{Shift mean}

Mean shift is another algorithm that is density based and mode seeking. It treats the data as a sample from an unknown probability density and looks for the local maxima (nodes) of this density, which will become the cluser centers. It uses a silding window (kernel) of radius h that is moved to the mean of the points inside the windows looking for the highest density. The windows are moved untill they converge, if they are really near they are put together inside the same cluster. 

\vspace{10pt}

Algorithm:
\begin{itemize}
    \item Create the kernel
    \item At every iteration move the window towards the most dense region within itself
    \item Repeat untill the movements are negligible
    \item Do this for every points untill we have clear saparations
\end{itemize}

The main characteristics are that all the data points will converge to the same cluster every time and that we have an attraction basin (a region where all the starting points will converge to a certain cluster).

\vspace{10pt}

Strengths

\begin{itemize}
    \item Can identify different shapes
    \item Can leave outliers as noise
    \item We dont't have to choose the number of clusters
    \item Good for small and medium datasets
    \item Versatile even though it can struggle in high dimensions
\end{itemize}

Weaknesses

\begin{itemize}
    \item Struggles with varying densities
    \item Sensitive to radius choice
    \item The radius is critical and difficult to set
    \item Expensive for large datasets
    \item Suffers from the curse of dimensionality
\end{itemize}

\subsubsection{Self-organizing maps}

With SOMs we want to train a unsupervised neural network where the inputs are connected to a matrix of neurons. Each neuron is a vector in the input space that during the training is pulled in the position of the input data dragging with them their neighbours in the output space. The map aims to pass between the data patterns. Input patterns are then compared with the neurons and the closest is considered the winner. The winner is updated and so its neighbors. The difference between data and neurons is called quantization error.

\vspace{10pt}

Algorithm

\begin{itemize}
    \item \textbf{Step 0:} Randomly initialize the weights $w_{ij}$.
    \item Set the neighborhood topological parameters.
    \item Set the learning rate $\alpha$.

    \item \textbf{Step 1:} While stop condition is false:
    \begin{itemize}
        \item \textbf{Step 2:} For each input vector $\mathbf{x}$:
        \begin{itemize}
            \item \textbf{Step 3:} For each unit $j$, compute  
            \[
                D(j) = \sum_i (w_{ij} - x_i)^2
            \]
            \item \textbf{Step 4:} Find the unit that minimizes $D(j)$ (Best Matching Unit).
            \item \textbf{Step 5:} For each unit $j$ in the neighborhood and for all $i$:
            \[
                w_{ij}^{new} = w_{ij}^{old} + \alpha \left(x_i - w_{ij}^{old}\right)
            \]
        \end{itemize}

        \item \textbf{Step 6:} Update the learning rate $\alpha$.
        \item \textbf{Step 7:} Reduce the radius of the topological neighborhood.
    \end{itemize}
\end{itemize}


\vspace{10pt}

From our SOM we get as output

\begin{itemize}
    \item U-matrices
    \item Component planes
    \item Hit plots
\end{itemize}

\subsection{Data Clustering}

Another problem is understanding how many clusters we want. One way is to create many cases with different K and choose the best. 
The choice should be done via
\begin{enumerate}
    \item Elbow curve method to lower intra cluster variability, we use a plot of the objective function value and the number of clusters
    \item Silhouette analysis, how similar are the points inside the clusters
    \item Evaluation of the profile of the cluster
    \item Operational considerations, based on business environment
\end{enumerate}

\vspace{10pt}

As said the shape of the clusters can be tricky for k-means, since is based on round shaped clusters. We can solve this by using k-medoids.

\vspace{10pt}

Another method to evaluate the clustering quality is the DBI. It focus on how compact the clusters are internally and how separated are from each other. We aim at lower values.
DBI is made by 3 key components: Scatter $S_i$ (average distance for the points in the cluster from the centroid), Separation $M_{ij}$ (distance between the centroids of 2 clusters) and similarity ratio $R_{ij}$ (how similar two clusters are, the lower the better).

To calculate the DBI

\begin{enumerate}
    \item Compute the centroid of each cluster
    \item Calculate $S_i$ for each cluster
    \item Calculate $M_{ij}$ for each pair of clusters
    \item Compute $R_{ij}$ for each cluster pair and select the maximum $R_{ij}$.
    \item DBI is given by the average of the $R_{ij}$
\end{enumerate}

\vspace{10pt}

We want to use elbow, Silhouette and DBI together and maximise them as much as possible to find the optimal k.
If they do not converge we need to choose based on the domain knowledge and Interpretability of our analysis.

It's very easy to interpret the elbow curve since we want to minimize the intra-cluster variability. The other 2 provides us for additional information (Silhouette how similar each point is to it's own cluster (higher is better) and DBI the average similarity ratio 
of each cluster with it's most similar one (lower is betetr))

The evaluation of the cluster profile is subjective dependent on the analyst.

Operational considerations are based on the business, usually we want a k small enough for developing a specific strategy,
a number if individuals large enough to be worth it to develop a specific startegy. Usually we can have an high initial k and grouping the clusters

\vspace{10pt}

Other important characteristics of the clusters are the shape and the density of the points. As said k-means can have problems with clusters of different size and density.
Also in this setting we are assuming that either an individual is part or not of a cluster, yes or no.

One variant is the k-medoids algorithm where each cluster is represented by one of the (real) points located near the center of the cluster.

\vspace{10pt}


\subsection{Association rules}

Another way to find patterns is the use of association rules, we want to extract compact patterns that describe subsets of data. Based on events that occur together to find rules on the form if X then Y. We get information about things that happen toegther. 

\vspace{10pt}

We can start with a table of purchased items to create a table of co-occurences with the number of times that any pair of products were purchased together. 

Usually association rules are expressed in the form of 

\begin{center}
\textit{"If the item A is part of an event, then the item B will also be part of the event X percent of time"}
\end{center}

The rules are not to be considered as causation but as association. All rules have an antecedent and a consequent \ra if a customer buys shoes, then 10\% of the time will also buy socks.

\vspace{10pt}

To evaluate rules we usually use

\begin{itemize}
    \item Confidence \ra How strong is the implication, the conditional probability that C occurs given the occurence of A
    \item Consequent support \ra Consequent transactions over the total number of transactions
    \item Support \ra How frequently the combination occurs in the database
    \item Lift \ra how much more often A and B occur together than would be expected if they were statistically independent, how strong the correlation is beyond random chance
\end{itemize}

\begin{tikzpicture}[every node/.style={font=\small}]
  % Outer universe U
  \draw[thick] (0,0) ellipse (3cm and 2cm);
  \node[above] at (0,2) {$U$};

  % A (left) and C (right) with opacity to show overlap clearly
  \filldraw[fill=blue!40, draw=black, opacity=0.5] (-0.9,0) ellipse (1.6cm and 1.2cm); % A
  \filldraw[fill=red!40,  draw=black, opacity=0.5] (0.9,0)  ellipse (1.6cm and 1.2cm); % C

  % Labels
  \node at (-1.5, 0.7) {$A$};
  \node at ( 1.5, 0.7) {$C$};
  \node at ( 0, -0.2) {$B$}; % intersection visible

\end{tikzpicture}


\begin{itemize}
    \item Confidence \ra $\frac{B}{A} = P(C \cap A)$
    \item Consequent support \ra $\frac{C}{U}$
    \item Support \ra $\frac{B}{U} = P(A \cap C)$
    \item Lift \ra $\frac{\text{Confidence}}{\text{Consequent support}} = \frac{P(C \cap A)}{P(C)}$
\end{itemize}


\vspace{10pt}

How to read 

\begin{enumerate}
    \item A lift of 2 means that two events will occur together twice as often as would be expected if their occurrences were independent
    \item A lift greater than 1 means a positive association
    \item A lift smaller than 1 means a negative association
    \item A lift equal to 1 means that that the two events are uncorrelated
    \item A credible rule should have good confidence, high support and a lift higher than 1
    \item High confidence and low support must be taken with caution
\end{enumerate}


\vspace{10pt}

Apriori algorithm for association rules

\begin{enumerate}
    \item Frequent itemset generation
    \item Create candidate itemsets
    \item Pruning infrequent itemsets
    \item Repeat the process
    \item Association rule generation
    \item Pruning rules
\end{enumerate}

\vspace{10pt}

Types of rules

\begin{itemize}
    \item Trivial 
    \item Inexplicable
    \item Actionable
    \item Rules obtained by promotions
\end{itemize}

We can also find product hierarchies. Remember that we need a lot of data about products and the customers.



\subsection{Dimensionality reduction}
Dimensionality reduction is the process of reducing the number of dimensions (features) in a dataset while retaining as much meaningful information as possible. It is used for visualization, computational efficiency and noise reduction. High-dimensional data often suffer from the curse of dimensionality.

Dimensionality reduction can be achieved through:

\begin{itemize}
    \item \textbf{Feature selection} → selecting a subset of relevant features
    \item \textbf{Feature extraction} → transforming data into a lower-dimensional representation
\end{itemize}

The goal is to preserve the structure of the original dataset when projecting it into a lower-dimensional space. Linear methods like PCA aim to preserve variance, while non-linear methods like t-SNE and UMAP focus on preserving neighborhood relations or the manifold structure.

High-dimensional data often lie on or near a lower-dimensional manifold. This motivates dimensionality reduction.

\vspace{10pt}

To recap:

\begin{itemize}
    \item Data redundancy
    \item Low intrinsic dimensionality
    \item Noise in high dimensions
    \item Manifold hypothesis
\end{itemize}

Dimensionality reduction can be obtained with:

\begin{itemize}
    \item Linear methods → PCA
    \item Non-linear methods → t-SNE or UMAP
\end{itemize}

PCA (Principal Component Analysis) is a linear method that identifies orthogonal directions (principal components) that maximize the projected variance. We keep those that explain most of the total variance.

t-SNE (t-Distributed Stochastic Neighbor Embedding) is a non-linear technique mainly used for visualization of high-dimensional clusters. It preserves local neighborhoods rather than global distances.

UMAP (Uniform Manifold Approximation and Projection) is also non-linear. It attempts to preserve both local and some global structure based on manifold and topological assumptions.

\vspace{10pt}

Dimensionality reduction is widely used in text mining to obtain dense vector representations. Classical representations include one-hot encoding, frequency vectors and TF-IDF (related to Zipf's law). More advanced embeddings include Word2Vec (CBOW or Skip-gram), which produce semantic clusters in vector space.

\subsubsection{PCA}

Principal Component Analysis aims at using an orthogonal transformation to convert a set of correlated variables into linearly uncorrelated principal components.  

If we have \(k\) features, PCA can produce up to \(k\) components (kernel PCA as well: it cannot exceed the rank of the data, which is \(\min(n, k)\)).  

Since PCA components are linear combinations of the original variables, PCA captures only linear relationships.

\vspace{10pt}

PCA steps:

\begin{enumerate}
    \item Standardize the data
    \item Compute the covariance matrix
    \item Perform eigendecomposition
    \item Sort eigenvectors by descending eigenvalues
    \item Select the principal components
    \item Create the projection matrix
    \item Transform the data
\end{enumerate}

\subsubsection{t-SNE}

t-SNE uses probabilistic similarity measures to map high-dimensional data to a lower-dimensional space (typically 2D or 3D). It preserves local neighborhood structure.  

\begin{itemize}
    \item Build conditional probabilities \(p_{j|i}\) over pairs of points in high-dimensional space  
    \item Build similar probabilities \(q_{ij}\) in low dimension using a Student-t distribution  
    \item Minimize the KL divergence between the two distributions  
\end{itemize}

\[
D_{KL}(P \parallel Q) = \sum_{i \ne j} p_{ij} \log\left(\frac{p_{ij}}{q_{ij}}\right)
\]

The Student-t distribution in low dimension helps alleviate the crowding problem.

\vspace{10pt}

The main parameter is \textbf{perplexity}, which controls the effective number of neighbors considered (typically between 5 and 50). It strongly influences the visualization.

\subsubsection{UMAP}

UMAP is a non-linear dimensionality reduction technique used for visualization and manifold learning, based on three assumptions:

\begin{itemize}
    \item Data is uniformly distributed on a manifold
    \item The manifold is locally connected
    \item The Riemannian metric is locally constant
\end{itemize}

With these assumptions, UMAP constructs a fuzzy topological representation of the data. The low-dimensional embedding is obtained by optimizing a layout that preserves this fuzzy structure.

\begin{itemize}
    \item Build the fuzzy representation of the manifold
    \item Optimize the low-dimensional embedding
\end{itemize}

The objective function is:

\[
L_{UMAP} = \sum_{i,j} \left[ p_{ij}\log(q_{ij}) + (1 - p_{ij})\log(1 - q_{ij}) \right]
\]

The first term represents \textbf{attractive forces}, the second \textbf{repulsive forces}.

Important parameters:
\begin{itemize}
    \item Number of nearest neighbors
    \item Minimum distance in the embedding
\end{itemize}