\section{Ethics for data science}


\subsection{Introduction}

\subsubsection{Law, moral and ethics}

Ethics as concept is old, but the application in data science is fairly recent. Ethics is really important in fields where the consequences are really important, like healthcare. 

\begin{itemize}
    \item \textbf{Ethics} \ra Moral principles, branch of philosophy that studies the basis of moral. Set of rules of conduct. Aims to direct the individual towards good practices. Individual perfecting. Based on the relationship of the individual with himself. Ethical rules are filtered by the individual, and it's conscience. Violation of ethical rules can lead to self-imposed sanctions. FOcussed on duties and obligations but not on rights.
    \item \textbf{Law} \ra Aims to regulate the essential relationships within a given society. Harmonious interactions and to solve conflicts between different people. Laws don't need individual conscience to be applied. Legal violations will lead to loss of liberties and life. Equal focus on duties and rights.
\end{itemize}

Laws waits for a conduct while ethics anticipate that conduct. Everyone can form an ethical opinion while legal ones are given by a judge. In ethical settings suspicion can affect someone while in legal ones people are innocent until proven guilty. 

\vspace{10pt}

Historically we saw innovations that lead to ethical principles and definitios (artificial life support created the legal definition of death or human genes discovery gave us the legal distinction of person vs thing).


\subsubsection{What is ethics}

We use ethics to make decisions related to choices that have moral dimensions. How one should act.

Ethics is made from two big branches

\begin{enumerate}
    \item Non-normative ethics \ra Studies what effectively the case is and not what ethically ought to be. We can have in this setting descriptive ethics (studies people beliefs about moralities) and meta-ethics (studies the meaning of moral propositions like right, obligations etc)
    \item Normative ethics \ra Studies the practical means of determining a moral course of action. It's applied ethics.
\end{enumerate}

We can also define the professional ethics branch, made of ethical gudielines set accepted by the majority of professionals. At a more granular level there exists also specific professional ethical rules. Some professions will adopt more detailed codes called Codes of ethical conduct. A public policy is a set of enforceable guidelines set by an official public body. 

\[\text{Public policy} = \text{Ethical principles} + \text{Empirical factors}\]



\subsubsection{Ethical dilemmas}

\begin{itemize}
    \item Psychologist and patient
    \item Cancer genes daughter
\end{itemize}


Ethical Dilemmas are circumstances in which ethical obligations demand or
appear to demand that a person adopt each of two (or more) alternative but
incompatible actions, such that the person cannot perform all the required
actions. Conflicts between ethical requirements and self-interest are
 practical dilemmas
and not necessarily
 ethical dilemmas

\vspace{10pt}

Tools of ethical decision making

\begin{itemize}
    \item \textbf{Basic ethics principles} \ra Aim to express general norms of common ethics (respect for autonomy, nonmaleficence, beneficence, justice)
    
    \item \textbf{Ethical rules} \ra More specific in content and more restricted in scope than principles. (however, both establish rights and obligations). Function as precise guides to action that direct us in each circumstance. They can be (substantive, authority and procedural)
    
    \item \textbf{Weighting and balancing} \ra Principles, rules, professional
obligations, and rights often
need to be weighted and
balanced.
Balancing is the process
of finding reasons to
support beliefs about
which ethical norms
should prevail. It is a process of practical astuteness, discriminating intelligence, and sympathetic
responsiveness. Necessary conditions are:
\begin{enumerate}
    \item  Good reasons can be offered to act on the overriding norm
rather than on the infringed
\item The moral objective justifying the infringement has a realistic
prospect of achievement

\item No morally preferable alternative actions are available
\item  The lowest level of infringement, commensurate with achieving
the primary goal of the action, has been selected
\item  Any negative effects of the infringement have been minimized
\item All affected parties have been treated impartially
\end{enumerate}

    \item \textbf{Virtues and emotions} \ra We want tho avoid ectremes situations like detachment or extreme attachment. In the absence of public and institutional constraints, partiality towards others is
ethically permissible and is the socially expected form of interaction. some necessary elements are discernment ( the ability to make fitting judgements and reach decisions without being influenced by extraneous factors), trustworthiness (Confident belief and reliance on the moral character and competence of another
person. Confidence that another will act with the right motives and in accordance with
ethical norms), integrity (Soundness, wholeness and integration of moral character. In a restricted sense means
fidelity in adherence to moral norms), concentiousness (Motivation to do what is right because it is right after trying with due diligence to
determine what is right) and compassion (Active regard for another´s welfare with an imaginative awareness and emotional
response of sympathy, tenderness and discomfort at another´s misfortune or suffering.)
\end{itemize}

\vspace{10pt}

Main ethical theories
\begin{itemize}
    \item Divine command
    \item Natural Law
    \item Virtue ethics
    \item Deontology
    \item Utilitarianism
\end{itemize}

A core point of ethical dilemmas is the disagreement that they can create given by different backgrounds, ideas or the quantity of information that everyone participating has. Disagreements may persist even among ethically committed persons.
Ethical disagreements may discourage persons who deal with practical problems but
should be no cause for skepticism about ethical thinking. 

\vspace{10pt}

Data ethics works on many levels based on the strenght of normative instruments. Remedies to violations can be
\begin{itemize}
    \item Prevention
    \item Detection/Investigation/Accusation
    \item Sanctions 
    \begin{itemize}
        \item Legal
        \item Disciplinary
        \item Professional
        \item Social
        \item Individual
    \end{itemize}
\end{itemize}

\subsection{History of data ethics}

Some case studies

\begin{itemize}
    \item Love is blind day
    \item Nurnmberg trial
    \item Tuskegee Experiment
    \item Milgram
    \item Rosenhan
    \item Tearoom trade
    \item Stanford prison 
\end{itemize}

Possible ways to approach digital revolution
\begin{itemize}
    \item Unchecked optimism
    \item Go backwards
    \item Accept with uneasiness
    \item Bend it towards justice and fairness
\end{itemize}

An important element for the last point is informed consent created after the Nuremberg code, it says that voluntary consent of the human subject is absolutely essential. Also the degree of risk should be proportional to the humanitarian impact.
Informed consent is made of 

\begin{itemize}
    \item Previous and precise aknowledgement of the situation
    \item Freedom from coercion
    \item Competence
    \item Flexibility
    \item Attention to vulnerable groups
\end{itemize}

Sometimes consent is impossible to obtain but it must be regularly evaluated.
Consent must be
\begin{itemize}
    \item Informed
    \item Voluntary
    \item Ethical 
\end{itemize}

The data ethics framewok is made of 

\begin{itemize}
    \item Overarching principles \ra Transparency, accountability, fairness
    \item Specific actions
\end{itemize}  


\subsection{Basics of organizational ethics}

Some snippets from Antonio Martins life. We learn our names and origins. 

\vspace{10pt}

Ethics is a set of principles, values and standards of conduct which guides individuals and organizations determining whats right and wrong. It involves making decisions and taking actions that are morally right considering the impact on stakeholders. Video on ethics.

\vspace{10pt}

We can consider organizational ethics as 
\begin{itemize}
    \item Trust and reputation
    \item Legal compliance
    \item Employee morale and engagement
\end{itemize}

We live in a society (professor said it). Joao from Leiria talks. 

\vspace{10pt}

Why companies need to be ethical

\begin{itemize}
    \item Building trust and reputation 
    \item Enhancing employee engagement and retention
    \item Mitigating risks and legal liabilities
\end{itemize}

Area where to use ethical approaches

\begin{itemize}
    \item Improving decision making and innovation
    \item Maintaining stakeholder relationships
    \item Demonstrating corporate social responsibility
\end{itemize}

When we talk about ethichs we talk about frameworks for organizational behaviour thus shaping how relationships work in an organization. Prioritizing ethical principles will make organizations creating sustainable and socially responsible business environment. When can also talk about ethical conduct, so how the company actually acts, but is not strictly morally right. 

\vspace{10pt}

Some ethical theories

\begin{enumerate}
    \item \textbf{Utilitarianism} \ra Maximise the happines of as many people as possible
    \item \textbf{Deontology} \ra We have some moral duties to follow
    \item \textbf{Virtue ethics} \ra Focus on virtues will lead to morally good lives
\end{enumerate}

Excercise about questions. 

\vspace{10pt}

Ethical stakeholders
\begin{itemize}
    \item Individuals
    \item Organizations
    \item Society
\end{itemize}

Key issues in data ethics 

\begin{itemize}
    \item Privacy
    \item Fairness
    \item Accountability
    \item Emerging challenges
\end{itemize}

\subsubsection{Privacy and confidentiality}

Personal data is a cornestone in digital economies, it's misuse can lead to privacy violations and heavy harm on all the actors invoved. Two important ways to save these data are a\textbf{Anonymization} and \textbf{differential privacy}. We define personal data as any information that relates to an identified or identifiable individual. Like basic identifiers, sensitive data, behavioural data and location data. 

Personal data have many values
\begin{itemize}
    \item Economic
    \item Strategic
    \item Technological advancement
    \item Societal insights
    \item Personal 
\end{itemize}

The main beneficiaries are
\begin{itemize}
    \item Businesses
    \item Governments
    \item Individuals
\end{itemize}

The risks are 
\begin{itemize}
    \item Data breaches
    \item Identity theft
    \item Surveillance and data erosion
    \item Exploitation
    \item Lack of consent
\end{itemize}

Two example of frameworks to save data privacy can be the \textbf{GDPR} and the \textbf{CCPA}.

\vspace{10pt}

To enforce \textbf{Personal Data Protection} we have some principles like:
\begin{itemize}
    \item Transparency
    \item Consent
    \item Data minimization \ra Masking, pseudonymization and generalization
    \item Security
    \item Accountability
\end{itemize}

Organizations should

\begin{itemize}
    \item Implement data privacy policies to inform users
    \item Use encryption to protect stored and transmitted data
    \item Adopt anonymization and differential privacy to minimize risks \ra Suppression, data swapping an K-anonimity. Noise addition, privacy budget, laplace and gaussian mechanisms
    \item Perform data audits to ensure compliance with legal standards
\end{itemize}

With data anonymization we can have some challenges to tackle:
\begin{itemize}
    \item Re-identification risk
    \item Utility vs privacy
    \item Linkability
\end{itemize}

Differential privacy has some key features
\begin{itemize}
    \item Mathematically proven privacy
    \item Aggregated insights
    \item Composable privacy
\end{itemize}

And some benefits

\begin{itemize}
    \item Strong privacy is guaranteed
    \item We can share data without violating regulations
    \item Prevention of re-identification
\end{itemize}

But also challenges
\begin{itemize}
    \item Balancing privacy and accuracy
    \item Complex to implement
    \item Cumulative noise
\end{itemize}

Differential privacy can be applied to
\begin{itemize}
    \item Census data
    \item Tech platforms
    \item Healthcare research
\end{itemize}

Individuals should

\begin{itemize}
    \item Manage consent
    \item Use strong passwords
    \item Be aware of terms
    \item Secure devices
    \item Minimize digital footprint
\end{itemize}

One real example of data breach was the Cambridge Analytica scandal

\subsubsection{Bias and fairness in statistical models}

A very big problem is the one of bias in data science. It can be caused by many elements and mitigated by
\begin{itemize}
    \item Audits
    \item Fairness metrics
    \item Diverse datasets
\end{itemize}

The bias can be created 
\begin{itemize}
    \item At data collection
    \item At model building
\end{itemize}

\subsubsection{Data sharing and collaboration}

Responsible data science must follow certain principles

\begin{itemize}
    \item Informed consent
    \item Data ownership
    \item FAIR \ra Promote findability, accessibility, interoperability and reusability
    \item Cross-border collaboration
\end{itemize}

\vspace{10pt}

\textbf{Informed consent}

\vspace{10pt}
Ensures that individuals understand

\begin{itemize}
    \item What data is being collected
    \item Why the data is being collected
    \item How the data will be shared, stored or processed
    \item Any associated risk
\end{itemize}

Informed consent empowers individuals and builds trust

\vspace{10pt}

Recommendations

\begin{itemize}
    \item Provide concise summaries
    \item Use interactive tools
    \item Allow individuals to opt-in than requiring opt-out actions
    \item Regularly update consent
\end{itemize}

Challenges

\begin{itemize}
    \item Information overload
    \item Power imbalances
    \item Digital platforms
\end{itemize}

\vspace{10pt}
\textbf{Data ownership}
\vspace{10pt}

Control, rights, and responsibilities individuals or organizations have over data.

\vspace{10pt}

Challenges
\begin{itemize}
    \item Ambiguity
    \item Cross-border data ownership
    \item Cloud services
\end{itemize}

Strategies
\begin{itemize}
    \item Legal contracts
    \item Data governance
    \item Empowering individuals
\end{itemize}

\vspace{10pt}
\textbf{FAIR principles}
\vspace{10pt}

Guidelines to ensure data is managed and shared ethically, promoting collaboration and innovation

\begin{itemize}
    \item \textbf{Findability}\ra Data should be easy to find
    \item \textbf{Accessibility} \ra Data should be easy to be retrieved
    \item \textbf{Interoperability} \ra Data should integrate easily with other datasets
    \item \textbf{Reusability} \ra Data should be well-documented for re-use
\end{itemize} 

Benefits
\begin{itemize}
    \item Ethical collaboration
    \item Efficiency
    \item Innovation
\end{itemize}
How to apply

\begin{enumerate}
    \item Use open repositories
    \item Apply machine-readable metadata
    \item Use open formats
    \item Document all the steps of data processing
\end{enumerate}

\vspace{10pt}
\textbf{Cross-border collaborations}
\vspace{10pt}
Sharing and processing data between institutions or individuals located
in different countries.

\vspace{10pt}

Opportunities

\begin{itemize}
    \item Global research
    \item Diverse insights
    \item Innovation
\end{itemize}

Challenges
\begin{itemize}
    \item Legal conflicts
    \item Data localization use
\end{itemize}

Ethical concerns
\begin{itemize}
    \item Ensuring informed 
    \item Avoiding exploitation
\end{itemize}

Technical barriers
\begin{itemize}
    \item Variations in data standards, infrastructure and formats
\end{itemize}

Best practices
\begin{itemize}
    \item Adopt global standards
    \item Draft clear agreements
    \item Secure data transfer
    \item Promote transparency
\end{itemize}


\subsubsection{Responsible use of predictive analytics and machine learning}

Predictive analytics can have different kinds of societal impacts, both on the negative and on the positive side

\begin{itemize}
    \item Positive \ra Enhance healthcare
    \item Negative \ra Over-reliance 
\end{itemize}


Over-reliance can cause
\begin{itemize}
    \item Loss of human oversight
    \item Automation bias
    \item Context ignorance
\end{itemize}

To fight over-reliance we can
\begin{itemize}
    \item Maintain a human-in-the-loop approach
    \item Algorithm audits
    \item Empower human judgement
\end{itemize}

Algorithms entail by design some flaws. Some targets that we want to reach are the maximization of fairness and transparency and the minimization of the effects of bias.

\vspace{10pt}

Some ethical dilemmas that arise in this context of automation are
\begin{itemize}
    \item The impact on employment
    \item Accountability of the decisions
    \item Risks for privacy
\end{itemize}

Some examples of success  stories in automation yadda yadda.



\subsubsection{Transparency and accountability}


Transparency and accountability are two very important elements if we want to ensure that our discipline is ethical, trustworthy and aligned with societal values. 

\vspace{10pt}

Explainability and transparency are important because we want to be able to understand why a decision was made and to go back if we find some bias. The challenges to take care of are

\begin{itemize}
    \item ML models usually are black boxes
    \item We lack data provenance
    \item Stakeholders might not be able to understand how the models work
\end{itemize}

Explainability can be obtained by
\begin{itemize}
    \item Add interpretability layers
    \item Add transparency during the collection of the data
    \item Communicate effectively with the stakeholders
\end{itemize}

To ensure accountability we can work with different frameworks

\begin{enumerate}
    \item Accountability mechanisms \ra audits and impact assessment
    \item Regulatory compliance \ra data protection laws, algorithmic accountability acts, cultural accountability
    \item Governance models \ra Ethics committee, cultural principles (ISO and NIST), stakeholder involvements, building trust, risk mitigation and regulatory compliance
\end{enumerate}

Practitioners can be recommended to do
\begin{itemize}
    \item Adopt explainability by design
    \item Conduct audits regularly
    \item Engage with stakeholders
    \item Establish clear governance policies
\end{itemize}




\subsubsection{Ethical decision-making process}


Now we see some courses of action to make ethical decisions.

\vspace{10pt}

Structured framework for individuals and organizations to evaluate ethical dilemmas and determine the most appropriate course of action

\begin{enumerate}
    \item Recognize the ethical issue
    \item Gather relevant information
    \item Evaluate alternative actions
    \item Make a decision
    \item Implement the decision
    \item Reflect on the outcome
\end{enumerate}

Kohlberg proposed a theory of moral development divided into 3 levels and 6 steps

\begin{enumerate}[label=\textbf{Level \Roman*}:]
    \item Pre-conventional morality
    \begin{enumerate}[label=\arabic*.]
        \item Obedience and punishment orientation $\rightarrow$ maintain personal safety
        \item Instrumental relativist orientation $\rightarrow$ gain rewards
    \end{enumerate}
    
    \item Conventional morality 
    \begin{enumerate}[label=\arabic*., resume] 
        \item Good interpersonal relationships $\rightarrow$ conforming to societal norms
        \item Maintaining social order $\rightarrow$ promote stability
    \end{enumerate}
    
    \item Post-conventional morality
    \begin{enumerate}[label=\arabic*., resume]
        \item Social contract and individual rights $\rightarrow$ greater good
        \item Universal ethical principles $\rightarrow$ conscience
    \end{enumerate}
\end{enumerate}

The application of the kinds of ethics can be done on certain parts of our previously defined framework

\begin{itemize}
    \item Utilitarianism \ra focus on the outcomes, we look for actions that mitigate harm and maximise benefits. We might justify morally questionable means.
    \item Deontology \ra we want to ensure that our moral principles are respected. We obtain clear and on-negotiable ethical boundaries. We can be stuck if we are too rigid, consequently we will disregard possible beneficial outcomes if we don't want to compromise on certain principles
    \item Virtue ethics \ra We accept implementations delays if it means that we are respecting ethical principles. This can lead to trust between the organization and the stakeholders. The decision-making process will be values-based but nonetheless subjective
\end{itemize}

In substance Utilitarianism is to evaluate the situation, Deontology to ensure that ethical boundaries are not crossed and virtue ethics to guide to actions aligned with the organization's identity and principles. 














\subsubsection{Promoting ethical culture}


Organizations should promote an ethical culture for different reasons, both towards their public image and towards a better way to organize towards themselves.

\begin{itemize}
    \item To communicate expectations
    \item To lead by example
    \item To foster an open communication
    \item To align incentives and rewards
    \item To provide training and support
\end{itemize}

In this setting we can identify three core elements
\begin{enumerate}
    \item Values \ra identify the core principles and beliefs that guide the decision-making. They are made of alignment, reinforcement and role modelling
    \item Norms \ra recognize unwritten rules with respect to what is acceptable or not. Key factors are socialization, peer influence and norm enforcement
    \item Leadership \ra assess the behavior and set the tone for the culture of the organization. Composed of tone at the top, decision-making practices and accountability
\end{enumerate}

Ethical values must be communicated, usually this is done in two main ways: 
\begin{itemize}
    \item By example
    \item By training and education
\end{itemize}

The development of ethical policies is done by
\begin{itemize}
    \item Whistleblower protection
    \item Ethical decision-making tools
\end{itemize}

The promotion of open communication and transparency is done by
\begin{itemize}
    \item Encouraging dialogue
    \item Transparency about decisions
    \item Enforcement of feedback mechanisms
\end{itemize}

To recognize ethical conduct
\begin{itemize}
    \item Performance evaluation
    \item Promotion of ethical leaders
\end{itemize}

Enforcement of ethical standards and accountability

\begin{itemize}
    \item Consistent enforcement
    \item Consequences for violations
    \item Ethical oversight
\end{itemize}

To improve and adapt over time
\begin{itemize}
    \item Regular evaluation
    \item Adaptation to change
    \item Learning from ethical lapses
\end{itemize}

Ethics in organizational process
\begin{itemize}
    \item Integration into decision-making
    \item Embedding in organizational structure
    \item Alignment with stakeholders expectations
\end{itemize}

\subsubsection{Ethical leadership}

Ethical leadership is a form of leadership characterized by integrity, honesty, transparency and commitment to ethical principles and values. Its basic characteristics are
\begin{itemize}
    \item Integrity
    \item Honesty and transparency
    \item Fairness and justice
    \item Empathy and compassion
    \item Accountability
    \item Courage
    \item Ethical decision-making
    \item Vision and purpose
    \item Ethical culture promotion
    \item Continuous learning and improvement
\end{itemize}

Challenges
\begin{itemize}
    \item Balancing competing interests
    \item Pressure to compromise ethics
    \item Ambiguity and uncertainty
    \item Resistance to change
    \item Ethical blind spots
    \item Ethical dilution in hierarchies
\end{itemize}

Strategies
\begin{itemize}
    \item Clarify ethical values and standards
    \item Lead by example
    \item Communicate ethical expectations
    \item Encourage open dialogue
    \item Provide ethical decision-making support
    \item Empower employees
    \item Foster a culture of trust and accountability
    \item Seek feedback and continuous improvement
    \item Address ethical lapses promptly and transparently
    \item Invest in ethical leadership development
\end{itemize}


\subsection{Cybersecurity}

We check if we have been hacked. F for who has a red mark.

\vspace{10pt}

\subsubsection{Basic definitions}

Cyber comes from the Greek word Kubernetes, meaning governor. The term cybernetic was coined by Norbert Wiener. Cyber was then put at the start of many words to relate them to the digital world. Another term is the cyberspace, a word used to describe the whole virtual world and its environments. 

\vspace{10pt}

In the cyberspace we can have both art and criminal activities, a very big space.

We define Cybersecurity as the ability to protect or defend the use of cyberspace from cyberattacks. 

\vspace{10pt}

Now that information is not stored on only physical devices we have to protect it in new ways since it can be accessed with many more devices. 

\vspace{10pt}

The three key objectives (CIA) are

\begin{enumerate}
    \item Confidentiality
    \item Integrity
    \item Availability
\end{enumerate}

We can define between cyber and information security. The first is for every kind of information stored in cyber systems while the second for information in general.



\subsubsection{The evolution of Cybersecurity}

The evolution of Cybersecurity followed the following steps

\begin{itemize}
    \item Antivirus
    \item Firewalls and encryption
    \item SOCs and Endpoint detection
    \item Zero trust and AI-Augmented defense
\end{itemize}

The first cyberattack was made in 1988 by Robert Morris. Bro talks about conspiracy theories. 

\subsubsection{Fundamentals of cyber risks and security management}

\subsubsection{Cybersecurity management}

\subsubsection{Cybercrime}

\subsubsection{Ethical issues in Cybersecurity}























