\section{Statistic for data science}

\subsection{Theory}

\subsubsection{Data exploration and basic concepts}

Statistics is the approach of answering questions using data. We aim to find rules that are true for a certain population basing our findings on a smaller subset of the mentioned population (sample)

\vspace{10pt}

\textbf{Unit} $\xrightarrow{}$ Basic objrcts on which the data are collected

\textbf{Variable} $\xrightarrow{}$ Characteristics of units that can take different values

\vspace{10pt}

We can have different types of variables
\vspace{10pt}

\textbf{Categorical}
\begin{itemize}
    \item Nominal
    \item Ordinal
\end{itemize}

\textbf{Quantitative}
\begin{itemize}
    \item Discrete
    \item Continuous
\end{itemize}

Our questions involve large \textbf{populations}. Since we can't measure all of the units of them we usually take a subset called \textbf{sample}. Values that concern the sample are called \textbf{statistics} while the ones that concern the whole population are called \textbf{parameters}. 

\vspace{10pt}

We want to estimate in a precise way those parameters basing the computation on the sample.

One of the problems regarding the samples is that those can be biased. To avoid that we want to get a sample that is representative of the population. One way is the use of \textbf{simple random sampling method}.

\textbf{Describing and visualizing data}

We have different ways to describe our data, they depend on multiple factors.

Categorical variables can be described by a proportion



\[
p = \frac{\text{Number in the category}}{\text{Total number}}
\]

To assess how many units share a certain value we can use a \textbf{Bar chart}

\begin{tikzpicture}
  \begin{axis}[
      ybar,
      bar width=15pt,
      xlabel={Categories},
      ylabel={Values},
      symbolic x coords={A,B,C,D},
      xtick=data,
      nodes near coords, % mostra i valori sopra le barre
      ymin=0,
  ]
    \addplot coordinates {(A,3) (B,7) (C,5) (D,2)};
  \end{axis}
\end{tikzpicture}

\vspace{10pt}

One important concept in statistics is central tendency, which refers to finding the center of our data. It also helps us understand how the other values are distributed around that center. Quantitative variables can be summarized using measures of central tendency.

\vspace{10pt}

\textbf{Mean} $\xrightarrow{}$ Sum of the data values divided by the number of the values. For the mean of the sample we use $\bar{x}$ and for the one of the population $\mu$. The formula is

\[
\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i
\]

\vspace{10pt}

\textbf{Median $\xrightarrow{}$} The center of sample after it was ordered from smallest to largest. Better than mean if we have outliers.

\vspace{10pt}

\textbf{Standard deviation} $\xrightarrow{}$ It describes how the units are distributed from the mean. We can square it to get the variance. For the sample we call it $s$, for the population we use $\sigma$. 

\[
s = \sqrt{\frac{1}{n-1} \sum_{i=1}^n (x_i - \overline{x})^2}
\]

Remember that both $\bar{x}$ and $s$ (or $s^2$) are relative to the sample. They are just an estimation that might be correct if we have a big enough sample.

\vspace{10pt}

\textbf{Percentile} $\xrightarrow{}$ Is a measure that tells us the percentage of the observations that fall at or below it. At $75\%$ we will find on the left the 75\% of the observations and on the right the 25\%.

\vspace{10pt}


\textbf{Quartiles} $\xrightarrow{}$ Basically the same as the percentiles but now we are dividing the sample in 4 parts. $Q_1$ is the first quartile, we will find on the left 25\% of the observations and the rest on the right. $Q_2$ is the median.

\[
Q_1 = \text{25th percentile}, \quad
Q_2 = \text{median (50th percentile)}, \quad
Q_3 = \text{75th percentile}
\]

\[
\text{Interquartile Range (IQR)} = Q_3 - Q_1
\]

\[
\text{Lower fence} = Q_1 - 1.5 \cdot \text{IQR}, \quad
\text{Upper fence} = Q_3 + 1.5 \cdot \text{IQR}
\]

\[
\text{Outliers} = \text{values outside } [\text{Lower fence}, \text{Upper fence}]
\]

\vspace{10pt}

We can visualize those statistics using a \textbf{Boxplot}

\begin{tikzpicture}
\begin{axis}[
    boxplot/draw direction=y,
    ylabel={Values},
    xtick={1},
    xticklabels={Sample Data},
    width=8cm,
    height=6cm
]

% boxplot principale
\addplot+[
    boxplot prepared={
        median=50,
        upper quartile=60,
        lower quartile=40,
        upper whisker=80,
        lower whisker=20
    },
] coordinates {};

% outlier come punti separati
\addplot+[
    only marks,
    mark=*,
    mark options={red},
] coordinates {(1,10) (1,15) (1,90) (1,95)};

\end{axis}
\end{tikzpicture}



\vspace{10pt}

\textbf{Correlation} $\xrightarrow{}$ Describes the strength and direction between two quantitative variables. For the sample we use $r$, for the population we use $\rho$.

\vspace{10pt}

Correlation has many properties:
\begin{enumerate}
    \item $\rho$ is bounded in [-1, 1]
    \item $\rho$ > 0 means positive association, $\rho$ < 0 negative, $\rho$ = 0 no relationship
    \item The bigger |$\rho$| the stronger the correlation
    \item The sign tells the direction
    \item $\rho$ is unit free
    \item $\rho$(X, Y) = $\rho$(Y, X)
    
\end{enumerate}

To visualize correlation we use a \textbf{Scatterplot}


\begin{tikzpicture}
  \begin{axis}[
      xlabel={X-axis},
      ylabel={Y-axis},
      title={Generic Scatter Plot},
      width=10cm,
      height=7cm,
      grid=major,
  ]
    % Dati del scatter plot
    \addplot[
      only marks,
      mark=*,
      mark size=2pt,
      color=blue
    ] coordinates {
        (1,2) (2,3) (3,5) (4,4) (5,7) (6,6)
        (1.5,2.5) (2.5,3.5) (3.5,4.5) (4.5,5.5)
        (1.2,2.8) (2.2,3.2) (3.2,5.1) (4.2,4.3)
        (5.1,7.2) (6.3,6.1) (1.7,2.9) (2.8,3.8)
        (3.8,4.2) (4.8,5.1) (5.5,7.3) (6.1,6.5)
    };
  \end{axis}
  
\end{tikzpicture}

\hrule

\subsubsection{Statistical Distributions}

\vspace{10pt}

Since random variables can take certain values with certain probabilities we would like to describe them somehow. The collection of those probability is called probability distribution. This can describe the probability of each event happening. Recall that the sum of the probabilities is 1.

\vspace{10pt}

\textbf{CDF} -> the \textbf{cumulative distribution function} gives me the probability that a random variable is less than or equal to a given value. 

\begin{equation}
    F(x) = P(X \le x) \quad \text{for all } x
\end{equation}


\textbf{PMF/PDF} -> the \textbf{probability mass function} or \textbf{probability density function} (for discrete or continuous case respectively) will give us the probability of a certain event happening

\begin{equation}
    f(x) = P(X = x), \quad \text{for all } x
\end{equation}

\begin{equation}
F(x) = \int_{-\infty}^{x} f(t)\,dt, \quad \text{for all } x
\end{equation}

For the continuous case we calculate the integral between the 2 values that we care 

\vspace{10pt}

\textbf{Discrete distributions}

If we can count the sample space we say that our random variable is discrete

Recall

\begin{equation}
    Var[X] = E[X^2] - E[X]^2
\end{equation}

\vspace{20pt}

\textbf{Bernoulli} -> the probability that a certain event will happen

\begin{equation}
    X \sim Ber(p)
\end{equation}
\begin{equation}
    P(X=1) = p = 1 -P(X=0) = 1-q
\end{equation}

With PMF

\begin{equation}
    f(x) = p^x(1-p)^x \quad \text{for x} \in \{0, 1\}
\end{equation}

Where

\begin{itemize}
    \item p -> positive probability
    \item q -> negative probability, 1-p basically
\end{itemize}

And with
\begin{itemize}
    \item E[X] = p
    \item Var[X] = pq = p(1-p)
\end{itemize}


\textbf{Binomial} -> probability of getting a certain number of getting n independent Bernoulli outcome

\begin{equation}
    X \sim Bin(n, p)
\end{equation}

\begin{equation}
    P(X=x) = \binom{n}{x}p^x(1-p)^{n-x}
\end{equation}

Where
\begin{itemize}
    \item n -> number of trials
    \item p -> positive probability
    \item x -> successes that we want to calculate
\end{itemize}

Recall that

\begin{equation}
    \binom{n}{x} = \frac{n!}{x!(n-x)!}
\end{equation}

And with
\begin{itemize}
    \item E[X] = np
    \item Var[X] = np(1-p)
\end{itemize}


\textbf{Poisson} -> probability of getting a certain number of events during a fixed interval. We need to know the probability of the event happening.

\begin{equation}
    X \sim Poi(\lambda)
\end{equation}

\begin{equation}
    f(x) = P(X=x) = \frac{\lambda^xe^{-\lambda}}{x!}
\end{equation}

Where

\begin{itemize}
    \item e -> Euler number
    \item x -> number of successes
    \item $\lambda $ -> known mean rate
\end{itemize}

With
\begin{itemize}
    \item E[X] = Var[X] = $\lambda$
\end{itemize}


\textbf{Continuous distributions}

\vspace{10pt}


\textbf{Normal distribution} -> If we want to describe a series of random events that have a tendency of being near a certain mean.

\begin{equation}
    X \sim N(\mu, \sigma^2
\end{equation}

\begin{equation}
    \phi (x) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp{\{-\frac{1}{2\sigma^2} (x-\mu)^2\}}
\end{equation}

\begin{itemize}
    \item E[X] = $\mu$
    \item Var[X] = $\sigma^2$
\end{itemize}

It has a bell shape and if the parameter are $\mu$ = 0 and $\sigma^2$ = 1 we say that is a standard normal distribution known as Z distribution. $\sigma^2$ will tell us how thin or wide is the distribution, so how variable it is, and so how unpredictable. 

Kurtosis -> Measure of how heavy the tails are, how frequent are the outliers

\vspace{10pt}

\vspace{10pt}

\textbf{Chi squared} -> If we have n independent X variables that are distributed with the Z distribution if we square and sum them we say that this quantity follows a Chi-squared distribution

\begin{equation}
    V=X_1^2+X_2^2+...+X_n^2 \sim\chi_{(n)}^2, n>0
\end{equation}

With
\begin{itemize}
    \item E[V] = n
    \item Var[V] = 2n
\end{itemize}

The n parameter is called \textbf{degrees of freedom}. The shape is asymmetric

\vspace{10pt}


\textbf{Student’s t distribution} $\rightarrow$  
If we have a random variable $Z$ distributed as a standard normal distribution and another independent variable $V$ following a Chi-squared distribution with $n$ degrees of freedom, then the following ratio defines a random variable that follows a t-distribution:

\begin{equation}
    T = \frac{Z}{\sqrt{V / n}} \sim t_{(n)}, \quad n > 0
\end{equation}

With
\begin{itemize}
    \item $E[T] = 0$ \quad (for $n > 1$)
    \item $Var[T] = \dfrac{n}{n - 2}$ \quad (for $n > 2$)
\end{itemize}

The parameter $n$ is called \textbf{degrees of freedom}.  
The distribution is symmetric around zero and has heavier tails than the normal distribution.  
As $n \to \infty$, the t-distribution approaches the standard normal distribution.

\vspace{10pt}


\textbf{Snedecor’s F distribution} $\rightarrow$  
If we have two independent random variables $U$ and $V$ such that  
$U \sim \chi^2_{(n_1)}$ and $V \sim \chi^2_{(n_2)}$,  
then the following ratio defines a random variable that follows an F-distribution:

\begin{equation}
    F = \frac{(U / n_1)}{(V / n_2)} \sim F_{(n_1, n_2)}, \quad n_1, n_2 > 0
\end{equation}

With
\begin{itemize}
    \item $E[F] = \dfrac{n_2}{n_2 - 2}$ \quad (for $n_2 > 2$)
    \item $Var[F] = \dfrac{2n_2^2 (n_1 + n_2 - 2)}{n_1 (n_2 - 2)^2 (n_2 - 4)}$ \quad (for $n_2 > 4$)
\end{itemize}

The parameters $n_1$ and $n_2$ are called \textbf{degrees of freedom} of the numerator and denominator, respectively.  
The distribution is asymmetric and always non-negative.  
It is mainly used to compare variances, especially in the context of ANOVA (Analysis of Variance) and regression models.


\vspace{10pt}

To summarize

\begin{table}[h]
    \centering
    \begin{tabular}{|c|c|c|c|} % '|' crea linee verticali
        \hline % linea orizzontale in alto
        \textbf{Distribution} & \textbf{PDF}  & \textbf{CDF}  & \textbf{Quantile} \\
        \hline % linea sotto l'intestazione
        Normal & dnorm & pnorm & qnorm \\
        \hline
        Chi-Squared & dchisq & pchisq & qchisq \\
        \hline
        Student-t & dt & pt & qt \\
        \hline
        F & df & pf & qf \\
        \hline % linea in fondo
    \end{tabular}
    \label{tab:distributions}
\end{table}

\vspace{10pt}

\hrule

\vspace{10pt}



\subsubsection{Point Estimation}

Suppose that we have a population distributed with certain parameters. Like $\mu$ as mean or p as proportion.
Since we can't check the full population (usually), we start from a sample and we try to see if the estimate can be good enough.

\vspace{10pt}

We have two main ways
\begin{itemize}
    \item Method of moments
    \item MLE
\end{itemize}

\vspace{10pt}

To start we need to define the \textbf{Parameter space}, the range of possible values of the parameter $\theta$ inside the parameter space $\Omega$.
Let $X_1, X_2, ..., X_n$ be n random variables that can take a value $x_1, x_2, ..., x_n$.
The parameter space is bounded.

\vspace{10pt}

To estimate the $\theta$ we use a point estimator. For example the mean is a point estimator of the $\mu$.

\begin{equation}
    \bar{X} = \frac{1}{n} \Sigma X_i
\end{equation}

We define the statistic T($X_1, ..., X_n$) as point estimator to get the parameter estimation.
If T($x_1, ..., x_n$) we get the estimate.


\vspace{10pt}

The first way is the \textbf{Method of moments}, we equate sample moments with theoretical moments.

\begin{itemize}
    \item E($X^k$) is the kth theoretical moment of the distribution, k=1,2,3\dots.
    \item $m_k$ = $ \frac{1}{n} \Sigma X_i^k$
\end{itemize}

We equate the first k sample moments to the corresponding k population moments until we have as many equations as se have
parameters, then we solve the resultant system.

\vspace{10pt}

The second way is the so called \textbf{Maximum Likelihood Estimator}. We start from the
assumption that in our sample the values follow a distribution based on an unknown 
parameter $\theta$. We want to find a good point estimator and estimate for $\theta$ using
the data from our sample. The idea is that we want to find the parameter that maximise
the probability of getting those values. If $X_1, ..., X_2$ are identically independent 
distributed  from a population with PDF f(x | $\theta_1, ..., \theta_k$) the likelihood 
function is defined by

\begin{equation}
    L(\theta|x) = L(\theta_1, ..., \theta_k | x_1, ..., x_k) = \Pi f(x_i|\theta_1, ..., \theta_k)
\end{equation}

After we take the log of the likelihood to obtain the log-likelihood function. This is because the logarithm is
a strictly increasing function so maximising the log of the likelihood is like maximising
the likelihood itself. The candidates are the values that put the derivative with respectively
to $\theta_i$ of the log-likelihood at 0. Also is easier to differentiate a log and working with sums

\begin{equation}
    logL(\theta|x) = \Sigma log(f(x_i | \theta_1, ..., \theta_k))
\end{equation}

\begin{equation}
    \frac{d}{d\theta_i} logL(\theta|\text{X}) = 0
\end{equation}

\vspace{10pt}

We want to assess how good are our estimates. To do it we want to look at the sampling distribution
that we receive. Our estimators have some properties.

\begin{itemize}
    \item \textbf{Unbiasedness} \ra An estimator T is unbiased iff E(T) = $\theta$ for every $\theta$ in $\Omega$. Usually the population mean is an ubiased estimator.
    \item \textbf{Efficiency} \ra We have two unbiased estimators T and T' for the same parameter $\theta$. T is more efficient than T' iff
    \begin{equation}
        Var(T) \le Var(T')
    \end{equation}
    T is the most efficient if the equation is true for any other unbiased estimator T'.
    This is called absolute efficiency, before it was only relative efficiency.
    To know that we have absolute efficiency we use the Frechet-Cramer-Rao inequality.

    Let $(X_1, X_2, \dots, X_n)$ be a sample from a population with pdf $f(x \mid \theta)$ and let
$T = T(X_1, X_2, \dots, X_n)$ be an unbiased estimator of $\theta$. Then

\[
\mathrm{Var}(T) \geq \frac{1}{n I(\theta)}
\]

where

\[
I(\theta) = \mathrm{E} \left[ \left( \frac{\partial \ln f(X \mid \theta)}{\partial \theta} \right)^2 \right]
= - \mathrm{E} \left[ \frac{\partial^2 \ln f(X \mid \theta)}{\partial \theta^2} \right]
\]

and $I(\theta)$ is the Fisher information.
\vspace{10pt}

This inequality allows us to find a lower limit to compare our variance with. 

\vspace{10pt}

If we want to compare relative efficiency, we can use the mean squared error (MSE). 

Let $T = T(X_1, \dots, X_n)$ be an unbiased estimator of the parameter $\theta$. The mean squared error is defined as

\begin{equation}
    \mathrm{MSE}(T) = \mathrm{E}[(T - \theta)^2].
\end{equation}

The estimator with the lower MSE is considered more efficient.

\item \textbf{Consistency} \ra Final property, it tells us that if the sample increases the precision of the estimator will also do that. An estimator $T_n$ is said to be consistent in quadratic mean iff 
\begin{equation}
    lim_{n \ra + \infty} \mathrm{E}[(T - \theta)^2] = 0
\end{equation}

\end{itemize}

\vspace{10pt}

\subsubsection{Hypothesis Testing}

\vspace{10pt}

Since we are doing inferential statistics we want to make inferences about our parameters.
As said we can't check the whole population but we need to consider a sample of it.
We use hypothesis testing to see if the value of a certain parameter can be ok with respect to the distribution of our population.

\vspace{10pt}

Hypothesis testing is done in 5 main steps

\begin{enumerate}
    \item Write the hypothesis
    \item Calculate the test statistic
    \item Determine the p-value
    \item Make a decision
    \item State a conclusion
\end{enumerate}

\vspace{10pt}

Let's delve into this list

\vspace{10pt}

The first step is to write the hypothesis, specifically the \textbf{Null} and the \textbf{Alternative}. The Null is denoted as $H_0$ while the Alternative as $H_1$.
The Null is always the statement that there is no difference in the population. The Alternative is always that we have differences.

\begin{table}[H]
\centering
\begin{tabular}{|p{3cm}|p{3.5cm}|p{3.5cm}|p{3.5cm}|}
\hline
\textbf{Research question} & Is the population mean \newline different from $\mu_0$ & Is the population mean \newline greater than $\mu_0$ & Is the population mean \newline less than $\mu_0$ \\
\hline
\textbf{Null Hypothesis ($H_0$)} & $\mu = \mu_0$ & $\mu = \mu_0$ & $\mu = \mu_0$ \\
\hline
\textbf{Alternative Hypothesis ($H_1$)} & $\mu \neq \mu_0$ & $\mu > \mu_0$ & $\mu < \mu_0$ \\
\hline
\textbf{Type of Hypothesis Test} & Two-tailed, \newline non-directional & Right-tailed, \newline directional & Left-tailed, \newline directional \\
\hline
\end{tabular}
\caption{Types of hypothesis tests for population mean, $mu_0$ is the hypothesized population mean}
\end{table}

\vspace{10pt}

The successive step is to calculate the test statistic. A value that tells us how much different are the values from the Null hypothesis.

For tests on the $\mu$ we use the following

\begin{equation*}
    t = \frac{\bar{X}-\mu_0}{\frac{\hat{\sigma}}{\sqrt{n}}} \sim t(n-1)
\end{equation*}


\vspace{10pt}

To estimate the sampling distribution for samples > 30 or with an unknown $\sigma$ we use a T-Student distribution with n-1 degrees of freedom.

If the standard deviation is known we use the normal distribution.

\vspace{10pt}

Now we want to calculate the p-value, or the probability that a population with the specified parameter would produce a sample statistic as extreme or more extreme to the one we observed in our sample. 
A test is significant if the p-value is less than or equal to the level of significance $\alpha$. If the p-value is greater we fail to reject the Null, if is less or equal we reject the Null.

To calculate a p-value for a T distribution on a two-tailed test we use

\begin{verbatim}
    2 * (1 - pt(test_statistic, df))
\end{verbatim}

\vspace{10pt}

Another approach is to use the \textbf{Critical region method}. In our case we get the critical region with

\begin{equation*}
    |t_obs| \geq t_{\frac{1-\alpha}{2};n-k-1}
\end{equation*}
 
Where the rightmost term is the quantile of the distribution. $\alpha$ is divided by 2 since is a two-tailed test.

If our test statistic is more extreme than the quantile we reject the Null.

To get the quantiles

\[
\texttt{qt(alpha/2, df)}
\]

\vspace{10pt}

We state the conclusion based on the results of the test.

\vspace{20pt}

Another possible kind of test is done to the proportions of the population.
We want to see if the proportion between different classes is realistic.

\begin{table}[h!]
\centering
\begin{tabular}{|p{3cm}|p{3.5cm}|p{3.5cm}|p{3.5cm}|}
\hline
\textbf{Research question} & Is the population \newline proportion different from $p_0$ & Is the population \newline proportion greater than $p_0$ & Is the population \newline proportion less than $p_0$ \\
\hline
\textbf{Null Hypothesis ($H_0$)} & $p = p_0$ & $p = p_0$ & $p = p_0$ \\
\hline
\textbf{Alternative Hypothesis ($H_1$)} & $p \neq p_0$ & $p > p_0$ & $p < p_0$ \\
\hline
\textbf{Type of Hypothesis Test} & Two-tailed, \newline non-directional & Right-tailed, \newline directional & Left-tailed, \newline directional \\
\hline
\end{tabular}
\caption{Types of hypothesis tests for population proportion, $p_0$ is our hypothesized proportion}
\end{table}

\vspace{10pt}

Usually we define the Null as p equal to a number

\vspace{10pt}

The test statistic is

\begin{equation*}
    Z_{obs} = \frac{\hat{p}- p_0}{\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}} \sim N(0,1)
\end{equation*}

With critical region

\begin{equation*}
    Z_{obs} \geq Z_{1 - \alpha}
\end{equation*}

Where 1 - $\alpha$ is the percentile of a standard normal distribution


\vspace{10pt}

When conducting a test we can make 2 possible decisions, either reject or fail to reject the null hypothesis. In most cases we can't be sure about our inference. If it's correct or not.


\vspace{10pt}

When we reject the Null we can make a \textbf{Type I} error. Also known as false positive.

When we fail to reject the Null we can make a \textbf{Type II} error. Also known as false negative.

\vspace{10pt}

The probability of rejecting a false Null hypothesis is known as power. 

\begin{equation*}
    \text{Power} = 1 - \beta
\end{equation*}

Where $\beta$ is the probability of committing a Type II error. $\alpha$ is the probability of making a type I error.



If we want to make \textbf{tests on two samples} we can follow the same idea.

For proportions

\[
\begin{cases}
    H_0 : p_A = p_B \\[4pt]
    H_1 : p_A > p_B
\end{cases}
\]
The test statistic will be

\[
Z_{obs} = \frac{\hat{p_A}-\hat{p_B}}{\sqrt{\frac{\hat{p_A}(1-\hat{p_A})}{n_A}+\frac{\hat{p_B}(1-\hat{p_B})}{n_B}}} \sim N(0,1)
\]


With critical region 
\[
Z_{obs} \ge Z_{1-\alpha}
\]


\vspace{10pt}

For means

\[
\begin{cases}
    H_0 : \mu_A = \mu_B \\[4pt]
    H_1 : \mu_A < \mu_B
\end{cases}
\]
The test statistic will be

\[
t_{\text{obs}} = 
\frac{\hat{\mu}_A - \hat{\mu}_B}{
\sqrt{
\frac{\hat{\sigma}_A^2}{n_A} +
\frac{\hat{\sigma}_B^2}{n_B}
}
}
\sim t(n_A + n_B - 2)
\]

If the samples have the same variance:

\[
t_{\text{obs}} = 
\frac{\hat{\mu}_A - \hat{\mu}_B}{
\hat{\sigma}
\sqrt{
\frac{1}{n_A} + \frac{1}{n_B}
}
}
\sim t(n_A + n_B - 2)
\]

With critical region:

\[
t_{\text{obs}} \le t_{\alpha; \, n_A + n_B - 2}
\]


Before testing for differences for means we should determine if the variances are equal


\[
\begin{cases}
    H_0 : \sigma_A^2 = \sigma_B^2 \\[4pt]
    H_1 : \sigma_A^2 \neq \sigma_B^2
\end{cases}
\]

With test statistic 

\[
F_{\text{obs}} = 
\frac{\hat{\sigma}_A^2}{\hat{\sigma}_B^2}
\sim F(n_A - 1, \, n_B - 1)
\]

With critical region

\[
F_{\text{obs}} \ge F_{\alpha; \, n_A - 1, \, n_B - 1}
\]





\subsubsection{Confidence Intervals}

Another way to perform statistical testing is via the use of Confidence Intervals. A CI is a range computed using sample statistic used to estimate a parameter of a population with a certain level of confidence.

The confidence is the difference between 1 and the alpha of the test statistic.


The general form of a confidence interval is 

\[
\bar{X} \pm t_{1-\frac{\alpha}{2}} (\frac{\hat{\sigma}}{\sqrt{n}})
\]

Basically with this form we want to estimate the population mean using the sample mean
, the percentile of the t distribution and the standard error.

\vspace{10pt}

When we use confidence Intervals we don't know if the parameter will fall inside the interval but we interpret it
as we are 95\% confident that the population parameter is between X and X'. Hence we get a range of acceptable estimates for the parameter while 
values that are outside won't be reasonable estimates. 

\vspace{10pt}

It's possible to notice that if we increase the sample size we will reduce the standard error, then we can get more precise estimates of our parameter. 


\subsubsection{Linear regression model}

Regression analysis is a disciplie that tries to uncover relationships between 2 or more variables.
This analysis is based on cross-sectional data, data that we obtain by random sampling from a population. Since we 
are having a random sampling process the covariance between the observations is zero. So for i$\ne$j

\[
Cov(y_i,y_j) = Cov(x_i, x_j) = Cov(y_i, x_j) = 0
\]

We want to calculate the causal effect of x on y, so how does y changes if x is changed and all other factors are held constant.

\vspace{10pt}

One of the most basic model that we can use is the simple linear regression model. We assume to have two variables y and x. We want to explain y in terms of x, so how y changes with changes in x.

To do that we need to confront some issues
\begin{itemize}
    \item How do we allow for other factors that can affect y
    \item What is the functional relationship between y and x 
    \item How to be sure to capture a ceteris paribus relationship between y and x 
\end{itemize}

The model follows this structure

\[
y = \beta_0 + \beta_1x + u
\]

In this equation u is the error term, a random disturbance in our population. It represents unobserved factors that aren't on x that can affect y.
The equation also addresses the issues of the functional relationship between x and y. Since if $\Delta$u=0 then x has a linear effect on y.

\[
\Delta y = \beta_1\Delta x
\]

The linearity part means that one unit change on x will have the same effect on y regardless of the initial value of x.

\vspace{10pt}

Another assumption is the conditional mean independence one. We say that unobserved factors do not contain information on explanatory variables.

\[
E[u|x]=0
\]

Given the previous notes about the simple linear regression model now we want to estimate the parameters.

Starting from 

\[
E[u \mid x] = 0 \;\Rightarrow\;
\begin{cases}
E[xu] = 0 \\
E[u] = 0
\end{cases}
\;\Leftrightarrow\;
\begin{cases}
E\big[x (y - \beta_0 - \beta_1 x)\big] = 0 \\
E\big[y - \beta_0 - \beta_1 x\big] = 0
\end{cases}
\]

Assuming that we have a random sample of n individuals and using the method of moments we get

\[
\frac{1}{n} \sum_{i=1}^{n} x_i \left( y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i \right) = 0
\tag{2}
\]

\[
\frac{1}{n} \sum_{i=1}^{n} \left( y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i \right) = 0
\]

That we can solve obtaining 

\[
\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}
\tag{4}
\]

\[
\hat{\beta}_1 =
\frac{\frac{1}{n} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}
{\frac{1}{n} \sum_{i=1}^{n} (x_i - \bar{x})^2}
\]


The estimates for the parameters are called OLS estimates and they are the best possible values for a simple linear regression model.

The error term can be calculated as

\[
\hat{u}_i = y_i - \hat{y}_i = y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i
\]

Those values are obtained by minimising the SSR by derivative with respect to both the $\beta_i$.

\vspace{10pt}

Recall

\[
\text{SST} = \sum_{i=1}^n (y_i - \bar{y})^2 = \text{SSR} + \text{SSE} \ra \text{Total sum of squares}
\]

\[
\text{SSR} = \sum_{i=1}^n (\hat{y}_i - \bar{y})^2 \ra \text{Regression sum of squares}
\]

\[
\text{SSE} = \sum_{i=1}^n (y_i - \hat{y}_i)^2 \ra \text{Error sum of squares}
\]


With the OLS method we will find the best solution every time. To evaluate it we can use the $R^2$, it measures the fraction of the sample variation in y that is explained by x.

\[
R^2 = \frac{SSE}{SST}=1-\frac{SSR}{SST}
\]


Sometimes linear dependencies are not enough so we can incorporate many non linearities into our models. 
Some of the most used are logarithmic transformations. 
\vspace{10pt}

We can start by defining the interpretations

\begin{itemize}
    \item Level-Level \ra No transformation is needed
    \item Log-Level \ra From exponential shape to linear 
       \item Log-Log \ra We have both logarithmic and exponential shapes and we transform them into linear
    \item Level-Log \ra From logarithmic shape to linear 
 
\end{itemize}

\begin{table}[h!]
\centering
\begin{tabular}{lccc}
\hline
\textbf{Model} & \textbf{Dep. Variable} & \textbf{Regressor} & \textbf{Interpretation of $\beta_1$} \\
\hline
Level--Level & $y$ & $x$ &
$\Delta E[y \mid x_i] = \beta_1 \,\Delta x$ \\[6pt]

Log--Level & $\log(y)$ & $x$ &
$\%\Delta E[y \mid x_i] \approx 100\,\beta_1\,\Delta x$ \\[6pt]

Log--Log & $\log(y)$ & $\log(x)$ &
$\%\Delta E[y \mid x_i] \approx \beta_1 \,\%\Delta x$ \\[6pt]

Level--Log & $y$ & $\log(x)$ &
$\Delta E[y \mid x_i] \approx \frac{\beta_1}{100}\,\%\Delta x$ \\
\hline
\end{tabular}
\end{table}

When we use simple regression we have the drawback that is very difficult to draw ceteris paribus conclusions about how x affects y.
The assumption that all the other factors affecting y are uncorrelated with x is often unrealistic. A solution is using multiple regression models. 
With this new setting is easier to fix certain values to get the actual importance of a certain x.

The general model with k explanatory variables has the following shape

\[
y_i = \beta_0 + \beta_1 x_{i1} + \cdots + \beta_k x_{ik} + u_i,
\qquad
\text{assuming } E[u_i \mid x_{i1},\ldots,x_{ik}] = 0.
\]

If we fix $x_2,...,x_k$ as $\Delta x_j = 0$ for j = 2,...,k we have

\[
\Delta\hat{y}_i = \hat{\beta}_i \Delta x_{i1}
\]

The residuals are obtained as 

\[
\hat{u}_i = y_i - \hat{y}_i
\]

And the following properties hold

\begin{enumerate}
    \item $\bar{y} = \bar{\hat{y}}$

    \item $\bar{y} = \hat{\beta}_0 
        + \hat{\beta}_1 \bar{x}_1 
        + \cdots 
        + \hat{\beta}_k \bar{x}_k$

    \item The sample covariance between each $x_j$, for $j = 1,\ldots,k$, 
          and $\hat{u}_i$ is zero. Hence, the sample covariance between 
          $\hat{y}_i$ and $\hat{u}_i$ is also zero.
\end{enumerate}

The multiple linear model follows certain assumptions

\begin{enumerate}

    \item \textbf{MLR1: Linear in Parameters} \\
    The model in the population can be written as
    \[
        y_i = \beta_0 + \beta_1 x_{i1} + \cdots + \beta_k x_{ik} + u_i
        = x_i' \beta + u_i .
    \]

    \item \textbf{MLR2: Random Sampling} \\
    We have a random sample of $n$ observations,
    \[
        \{(x_{i1}, x_{i2}, \ldots, x_{ik}, y_i)\}_{i=1}^n,
    \]
    following the population model in Assumption MLR1.

    \item \textbf{MLR3: No Perfect Collinearity} \\
    In the sample there are no exact linear relationships among the independent variables.  
    Put otherwise, the matrix $X$ has rank $k+1$.

    \item \textbf{MLR4: Zero Conditional Mean} \\
    Conditional on the entire matrix $X$, each error $u_i$ has zero mean:
    \[
        E[u_i \mid X] = 0.
    \]

    \item \textbf{MLR5: Homoskedasticity} \\
    The error $u$ has the same variance given any value of the explanatory variables:
    \[
        \operatorname{Var}(u \mid X) = \sigma^2 I_n,
    \]
    where $I_n$ denotes the $n$-dimensional identity matrix.
\end{enumerate}

\vspace{20pt}
\begin{theorem}[Unbiasedness of OLS]
Under assumptions MLR1--MLR4, the OLS estimators are unbiased estimators 
of the population parameters. We have:
\[
    E[\hat{\beta} \mid X] = \beta .
\]
\end{theorem}

\begin{theorem}[Variances of OLS Estimators]
Under assumptions MLR1--MLR5, the variance--covariance matrix of the OLS 
estimators is:
\[
    \operatorname{Var}(\hat{\beta} \mid X)
    = \sigma^{2}(X'X)^{-1}.
\]
\end{theorem}

\begin{theorem}[Gauss--Markov Theorem]
Under assumptions MLR1--MLR5, the OLS estimators are the Best Linear Unbiased 
Estimators (BLUE) of the regression coefficients. That is,
\[
    \operatorname{Var}(\hat{\beta}_j \mid X)
    \le
    \operatorname{Var}(\tilde{\beta}_j \mid X),
\]
for any estimator $\tilde{\beta}_j$ that is linear in $y$ and unbiased.  
In matrix form:
\[
    \operatorname{Var}(\hat{\beta} \mid X)
    \le
    \operatorname{Var}(\tilde{\beta} \mid X).
\]
\end{theorem}

\vspace{10pt}

Now that we have defined our assumptions and description of the linear models we want to make inferences on the population parameters. 
We can do that by constructing confidence intervals or testing possible values. We know that OLS estimators are random variables and we know their expected values and variances. But 
we want to know their distribution. In order to derive those distributions we need additional assumptions, for example that the errors are normally distributed.

\begin{enumerate}[resume]
    \item \textbf{MLR6: Normality} \\
    We have
    \[
    \textbf{u} | \textbf{X} \sim N(\textbf{0}, \sigma^2 \textbf{I})
    \]
\end{enumerate}

We know that the error term is a sum of many different unobserved factors that are normally distributed. Under normality assumption OLS is BLUE. If we can't assume normality we can use a large sample size.

\begin{itemize}
    \item Assumptions 1-5 \ra Gauss-Markov assumptions
    \item Assumptions 1-6 \ra Classical assumptions
\end{itemize}

\vspace{10pt}

\begin{theorem}[Normal Sampling Distribution of OLS]
Under Assumptions MLR1--MLR6 we have
\[
    \hat{\beta}_j \mid X \sim 
    N\!\left( 
        \beta_j,\;
        \operatorname{Var}\!\left( \hat{\beta}_j \mid X \right)
    \right).
\]
Therefore,
\[
    z = 
    \frac{
        \hat{\beta}_j - \beta_j
    }{
        \sqrt{\operatorname{Var}\!\left( \hat{\beta}_j \mid X \right)}
    }
    \sim N(0,1).
\]
\end{theorem}

\begin{theorem}[t-distribution for Standardized Estimators]
Under Assumptions MLR1--MLR6 we have
\[
    t = 
    \frac{
        \hat{\beta}_j - \beta_j
    }{
        \sqrt{\widehat{\operatorname{Var}}\!\left( \hat{\beta}_j \mid X \right)}
    }
    \sim t(n - k - 1).
\]
\end{theorem}

\vspace{10pt}

If we want to make inferences on $\beta$ we will have
\[
\begin{array}{l}
H_0 : \beta_j = a \\
H_1 : \beta_j \ne a
\end{array}
\]

We have two possibilities

\begin{itemize}
    \item If $|t_{obs}| > t_{\alpha/2}$ \ra Reject $H_0$ at $\alpha$ level
    \item If $|t_{obs}| \le t_{\alpha/2}$ \ra Do not reject $H_0$
\end{itemize}
where 
\[
t_{\text{obs}} =
\frac{\hat{\beta}_j - a}
{\sqrt{\widehat{\operatorname{Var}}\!\left( \hat{\beta}_j \right)}}
\]

\[
t_{\alpha/2} : \quad P(t > t_{\alpha/2}) = \frac{\alpha}{2}
\]

\textbf{Definition:} 
The p-value is the probability of obtaining a test statistic at least as extreme as the one actually observed, 
assuming that the null hypothesis is true.

\begin{itemize}
    \item Calculate the p-value for $H_0 : \beta_j = a$:
    \begin{itemize}
        \item $H_1 : \beta_j \ne a \implies \text{p-value} = 2 P\big(|t| > |t_{\text{obs}}| \mid H_0 \text{ is true}\big)$
        \item $H_1 : \beta_j > a \implies \text{p-value} = P\big(t > t_{\text{obs}} \mid H_0 \text{ is true}\big)$
        \item $H_1 : \beta_j < a \implies \text{p-value} = P\big(t < t_{\text{obs}} \mid H_0 \text{ is true}\big)$
    \end{itemize}
    
    \item For example, a p-value of $0.02$ indicates little evidence supporting $H_0$.
    
    \item At the 5\% significance level, you would reject the null hypothesis.
    
    \item \textbf{Rejection rule:}
    \begin{itemize}
        \item p-value $> \alpha \implies$ do not reject $H_0$ at the $\alpha \times 100\%$ level
        \item p-value $\le \alpha \implies$ reject $H_0$ at the $\alpha \times 100\%$ level
    \end{itemize}
\end{itemize}

Very often we want to test whether there is any relation between $x_j$ and the expected value of $y$, without imposing a sign on the partial effect a priori.

Thus, in most applications, our primary interest lies in testing the null hypothesis
\[
H_0 : \beta_j = 0.
\]

Here, $\beta_j$ measures the partial effect of $x_j$ on the expected value of $y$, after controlling for all other independent variables.  

If $\beta_j = 0$, this means that, once the other variables have been accounted for, $x_j$ has no effect on the expected value of $y$.

\vspace{10pt}

Suppose we want to test a single hypothesis involving more than one of the $\beta_j$.  

The null and alternative hypotheses are:

\[
H_0 : \beta_1 - \beta_2 = 0 \quad \text{vs.} \quad H_1 : \beta_1 - \beta_2 < 0
\]

Under the null, the corresponding $t$-statistic is

\[
t = \frac{\hat{\beta}_1 - \hat{\beta}_2}{\text{se}(\hat{\beta}_1 - \hat{\beta}_2)}.
\]

and 

\[
\text{se}(\hat{\beta}_1 - \hat{\beta}_2) 
= \sqrt{
    \widehat{\operatorname{Var}}(\hat{\beta}_1)
    + \widehat{\operatorname{Var}}(\hat{\beta}_2)
    - 2 \, \widehat{\operatorname{Cov}}(\hat{\beta}_1, \hat{\beta}_2)
}.
\]


















\subsection{Exercises}  

\begin{itemize}
    \item We have a N$\sim$(102, 8) find the area less than 120

    \begin{verbatim}
        pnorm(120, mean=102, sd=8, lower.tail=TRUE)
    \end{verbatim}

\item Same as above but we want the area more than 120

\begin{verbatim}
    pnorm(120, mean=102, sd=8, lower.tail=FALSE)
\end{verbatim}

\item Proportion of a Z distribution where the z score is between 0 and 1.75

\begin{verbatim}
    pnorm(1.75, mean=0, sd=1, lower.tail=TRUE) - 
    pnorm(0, mean=0, sd=1, lower.tail=TRUE)
\end{verbatim}

\item Proportion of a Z distribution more extreme than $\pm$2

\begin{verbatim}
    pnorm(2, mean=0, sd=1, lower.tail=False)*2
\end{verbatim}

\item Which z-scores separate middle 90\% and the outer 10\%

\begin{verbatim}
    qnorm(0.95, mean=0, sd=1, lower.tail=TRUE)
\end{verbatim}

\item We have a N$\sim$(102, 8), which value separates the top 10\%

\begin{verbatim}
    qnorm(0.90, mean=102, sd=8, lower.tail=TRUE) 
\end{verbatim}

\item Find the 95th percentile of a chi-squared with 7 df

\begin{verbatim}
    qchisq(0.95, df=7)
\end{verbatim}

\item Find 2.5th and 97.5 percentiles of a Student distribution with 5 df

\begin{verbatim}
    qt(0.95, df=5)

    or

    qt(c(0.25, 0.975), df=5)
\end{verbatim}

\item Finde 95th percetile of an F distribution with 12, 5 df

\begin{verbatim}
    qt(0.95, df1=12, df2=5)
\end{verbatim}




\subsection{Code}
    
\end{itemize}
